---
title: Schedule
layout: page
icon: "far fa-calendar-alt"
output: 
  md_document:
    toc: true
---



```{r btngrp, echo=FALSE}
knitr::raw_html(
'<div class="btn-group" role="group" align-items="center" aria-label="Basic example">
  <button type="button" class="btn btn-secondary" href="0-introduction-and-review">0 Intro</button>
  <button type="button" class="btn btn-secondary" href="1-model-accuracy">1 Model Checks</button>
  <button type="button" class="btn btn-secondary" href="2-regularization-smoothing-and-trees">2 Regression</button>
  <button type="button" class="btn btn-secondary" href="3-classification">3 Classification</button>
  <button type="button" class="btn btn-secondary" href="4-modern-techniques">4 Modern</button>
  <button type="button" class="btn btn-secondary" href="5-unsupervised-learning">5 Unsupervised</button>
</div>'
)
```

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
suppressPackageStartupMessages(library(tidyverse))
knitr::opts_chunk$set(echo=FALSE, cache=FALSE, autodep=FALSE, message=FALSE, warning=FALSE)
sched <- read_csv("assets/data/schedule.csv") %>% select(Date:Activity)
specials <- sched %>% filter(Activity != "Lab") %>% 
  mutate(start = Date, content = Activity, group = as.integer(as.factor(Activity)))
cols = RColorBrewer::brewer.pal(length(unique(sched$Module)), "Set1")
specials = specials %>% 
  mutate(style=paste0("background-color: ",cols)[Module+1])
```

```{r timeline, eval=FALSE}
ng = max(specials$group)
timevis::timevis(
  specials, groups = 
    data.frame(id=1:ng, content=rep("",ng)))
```



# 0 Introduction and Review
<a name="0intro"></a>

Reading 
: [ISLR] Ch 2.1, 2.2  and Chapter 3 (should be review)

Video lectures
: Lecture 0 (in class on Thursday)

```{r}
tab0 <- knitr::kable(select(filter(sched, Module==0), Date, Activity))
```

# 1 Model Accuracy
<a name="1model"></a>

Reading 
: [ISLR] Ch 2.2, 5.1, 6.1, 6.4

Video lectures
: Lecture 0 (in class on Thursday)

```{r}
knitr::kable(select(filter(sched, Module==1), Date, Activity))
```

## 2 Regularization, smoothing, and trees
<a name="2reg"></a>

Reading 
: [ISLR] Ch 6.2, 7.1&nbsp7.7.1, 8.1, 8.1.1, 8.1.3, 8.1.4

Video lectures
: Lecture 0 (in class on Thursday)

```{r}
knitr::kable(select(filter(sched, Module==2), Date, Activity))
```


## 3 Classification
<a name="3class"></a>

Reading 
: [ISLR] Ch 2.2.3, 4&nbsp4.5, 8.1.2

Video lectures
: Lecture 0 (in class on Thursday)

```{r}
knitr::kable(select(filter(sched, Module==3), Date, Activity))
```


## 4 Modern techniques
<a name="4modern"></a>

Reading 
: [ISLR] 5.2, 8.2

Video lectures
: Lecture 0 (in class on Thursday)

```{r}
knitr::kable(select(filter(sched, Module==4), Date, Activity))
```

## 5 Unsupervised learning
<a name="5unsup"></a>

Reading 
: [ISLR] 10

Video lectures
: Lecture 0 (in class on Thursday)

```{r}
knitr::kable(select(filter(sched, Module==5), Date, Activity))
```

## Midterm 5 and Final exam

TBA &nbsp During University Scheduled Final Exam Period
