---
title: "00 Programming"
author: 
  - "STAT 406"
  - "Daniel J. McDonald"
date: 'Last modified - `r Sys.Date()`'
output: pdf_document
---


## General advice

When writing R code (or any code), there are some important rules

1. Write script files (which you save) and source them. Don't do everything in the console.
2. Don't write anything more than once. This has three corollaries:
    a. If you are tempted to copy/paste, don't.
    b. Don't use _magic numbers_. Define all constants at the top of the script.
    c. Write functions.
3. The third is __very important__. Functions are easy to test. You give different inputs and check
whether the output is as expected. This helps catch mistakes.
4. There are two kinds of errors: syntax and function.  
    a. The first R can find (missing close parenthesis, wrong arguments, etc.  
    b. The second you can only catch by thorough testing (see the HW)
5. Don't use __magic numbers__. 
6. Use meaningful names. Don't do this:
```{r general-advice}
data("ChickWeight")
out <-lm(weight ~ Time + Chick + Diet, data = ChickWeight)
```
7. Comment things that aren't clear from the (meaningful) names
8. Comment long formulas that don't immediately make sense: 
```{r general-advice-2}
garbage = with(ChickWeight, 
               by(weight, Chick, 
                  function(x) (x^2 + 23) / length(x)
                  )
               ) ## WTF???
```

## Simple stuff

* Vectors:
```{r}
x <- c(1, 3, 4)
x
x[1]
x[-1]
```

* Matrices:
```{r}
x <- matrix(1:25, 5, 5)
x
x[1,]
x[,-1]
x[c(1,3),  2:3]
```


## Functions

Write lots of functions. I can't emphasize this enough.

```{r functions}
f <- function(arg1, arg2, arg3=12, ...){
  stuff <- arg1*arg3
  stuff2 <- stuff + arg2
  plot(arg1, stuff2, ...)
  return(stuff2)
}
x <- rnorm(100)
y1 <- f(x, 3, 15, col=2, pch=19)
f(x, 3)
```

## Outputs vs. Side effects

* Side effects are things a function does, outputs can be assigned to variables
* A good example is the `hist` function
* You have probably only seen the side effect which is to plot the histogram
```{r,fig.align='center'}
myHistogram <- hist(rnorm(1000))
```

## The output

```{r}
myHistogram
```

## Assignment

> What's up with `<-` and `=`?

* These two work mostly the same but not always.
* The code `<-` means to assign the stuff on the right to the name on the left:
```{r}
x <- 12
x; rm(x) # removing it so I can demonstrate 
```
This gives `x` the value `12`.

* Technically, this is the same as
```{r}
assign('x', 12)
x; rm(x)
```

## Versatility

```{r, echo=FALSE}
knitr::opts_chunk$set(error=TRUE)
```
* In that simple case `=` does the same thing. However, `<-` is more versatile. Consider:
```{r}
median(x = 1:10)
x
```
```{r}
median(x <- 1:10)
x
```

## General practice

* Many style guides say to __always__ use `<-`.
* If you use `<-`, you should put a space on both sides. This avoids issues like
```{r}
x< -3
```
when you meant
```{r}
x <- 3
```
* One reason to avoid `=` is due to confusion with logical operators like "Does x=1?"
```{r}
x=1
x==1
```

## Flow control

```{r flow-control}
x = 1
y = c(2,3,4,1,-1,0)
# bad  if(x=1) print(x)
if (x == 1) print(x)
y > x
any(y > x)
! x
all(y > x)
while(x < 4){
  print(x)
  x = x + 1
}
for (i in 1:4) print(x + i)
ifelse(any(y > x), 'yes', 'no')
```

## Some functions

```{r}
qpareto1 <- function(p, exponent, threshold) { 
  # gives the inverse cdf (quantile) of a Pareto distribution 
  # takes 3 parameters, exponent and threshold are for the distribution
  # p is the probability you want
  # See https://en.wikipedia.org/wiki/Pareto_distribution
  threshold * ( (1 - p) ^ (-1 / (exponent - 1))) # returns the last object
}

qpareto3 <- function(p, exponent, threshold, lower.tail=TRUE) {
  # add a *default* argument to decide whether we want the lower/upper tail
  if (!lower.tail) p <- 1 - p 
  q <- qpareto1(p, exponent, threshold)
  return(q) # explicitly say waht to return, this line is not required
}
```
```{r}
qpareto1(.4, 2, 2)
qpareto3(.4, 2, 2)
qpareto3(.4, 2, 2, FALSE)
qpareto3(.6, 2, 2)
```



## Traceback

```{r}
qpareto4 <- function(p, exponent, threshold, lower.tail = TRUE) {
  # just like qpareto 3, but now we check that our inputs make sense
  stopifnot(p >= 0, p <= 1, exponent > 1, threshold > 0) 
  q <- qpareto3(p, exponent, threshold, lower.tail)
  return(q)
}
rpareto <- function(n, exponent, threshold) { 
  # generates random numbers
  # Uses the "inverse cdf" method
  # https://en.wikipedia.org/wiki/Inverse_transform_sampling
  x <- vector(length = n)
  for (i in 1:n) 
    x[i] <- qpareto4(rnorm(1), exponent, threshold) # there's an error here
  x
}
```
```{r}
rpareto(10, 3, 1)
```

* Try `traceback` in RStudio

## Vectorizing

```{r}
rpareto <- function(n, exponent, threshold) {
  x <- double(n) # always preallocate space to fill in a loop
  for (i in 1:n) 
    x[i] <- qpareto4(p=runif(1), exponent, threshold) # this one is correct
  x
}
rpareto_better <- function(n, exponent, threshold) {
  qpareto4(runif(n), exponent, threshold)
  # we don't need to loop because these operations operate on the vector
}
system.time(rpareto(1e6, 2, 1))
system.time(rpareto_better(1e6, 2, 1))
```

## When might loops bad?


* The short answer is that `R` is not a __compiled__ language.
* This means that whenever you write a loop, `R` has to re-read all the code within the loop each iteration
* This is muy slow.
* The only thing slower, is if you don't preallocate.
* Remember that line `x <- vector(length=n)`?
* Without that line, `x` would get built within the loop, starting with length 1, then length 2, etc.
* Preallocation is the most important issue to address when writing loops.


## lapply vs. apply vs. sapply

* Many functions are __vectorized__, but not all.
* Arithmetic functions __are__
```{r}
1 + 1
c(1,2,3) + c(4,5,6)
c(1,2,3) + 1
```
* Some strange ones
```{r}
min(5:1,pi)
pmin(5:1,pi)
```

## The apply variants

* These try to do things where simple loops would suffice.
* `apply` is for matrices (or arrays). If you want to __apply__ a function along a dimension
```{r}
(mat <- matrix(1:100,10)) # the parenthesis makes the result print
sum(mat)
apply(mat,2,sum) # "applies" the function "sum" to each column (2nd dimension)
for(i in seq_len(ncol(mat))) sum(mat[ ,i]) # same in a loop
```

## lapply and sapply


* These work for lists
```{r}
(z <- list(a = 1:5, b = matrix(rnorm(10), 2), c = 25))
lapply(z, sum)
sapply(z, sum)
```

## lapply craziness

What does this do?

```{r}
sapply(lapply(1:10, rnorm), mean)
```

# Linear models

## Predict and Friends

* R has lots of functions for working with different sorts of 
predictive models.

* We should review how they work with `lm`, and how they generalize to other sorts of models.  

* We'll use the __Mobility__ data we saw earlier:

```{r}
mob <- readRDS("../data/mobility.RDS")
```

## Estimation Functions and Formulas


* To estimate a linear model in `R`: you use `lm`.  

```{r}
mob.lm1 <- lm(mob$Mobility ~ mob$Population +
                mob$Seg_racial + mob$Commute + mob$Income + mob$Gini)
```

* What `lm` returns is a complex object containing the estimated coefficients, the fitted values, a lot of diagnostic statistics, and a lot of information about exactly what work R did to do the estimation.  We will come back to some of this later.  

* The thing to focus on for now is the argument to `lm` in the
line of code above, which tells the function exactly what model to estimate 

* it __specifies__ the model.  The `R` jargon term for that sort of specification is that it is the **formula** of the model.


## The data argument

* While the line of code above works, it's not very elegant, because we have to
keep typing `mob$` over and over. 

* More abstractly, it runs specifying which
variables we want to use (and how we want to use them) together with telling R
where to look up the variables.  This gets annoying if we want to, say, compare
estimates of the same model on two different data sets (in this example,
perhaps from different years).  

* The solution is to separate the formula
from the data source:

```{r}
mob.lm2 <- lm(Mobility ~ Population + Seg_racial + Commute + Income + Gini, data=mob)
```

* The `data` argument tells `lm` to look up variable names appearing in the
formula (the first argument) in a dataframe called `mob`.  

* It therefore works
even if there aren't variables in our workspace called `Mobility`,
`Population`, etc., those just have to be column names in `mob`.  

* In addition
to being easier to write, read and re-use than our first effort, this format
works better when we use the model for prediction, as explained below.

## Transformations 


```{r}
mob.lm3 <- lm(
  Mobility ~ log(Population) + Seg_racial + Commute + Income + Gini, 
  data = mob
)
```

* Formulas are so important that `R` knows about them as a special data type. 

* They _look_ like ordinary strings, but they _act_ differently, so there are special
functions for converting strings (or potentially other things) to formulas, and
for manipulating them.  

* For instance, if we want to keep around the formula
with log-transformed population, we can do as follows:

```{r}
form.logpop <- "Mobility ~ log(Population) + Seg_racial + Commute + Income + Gini"
form.logpop <- as.formula(form.logpop)
mob.lm4 <- lm(form.logpop, data = mob)
```

## Why formulas?


* Being able to turn strings into formulas is very convenient if we want to try out a bunch of different model specifications, because R has lots of tools for building strings according to regular patterns, and then we can turn all those
into formulas.  



* If we have already estimated a model and want the formula it used as the specification, we can extract that with the `formula` function:

```{r}
formula(mob.lm3)
formula(mob.lm3) == form.logpop
```

## Extracting Coefficients, Confidence Intervals, Fitted Values, Residuals, etc.

If we want the coefficients of a model we've estimated, we can get that
with the `coefficients` function:

```{r}
coefficients(mob.lm3)
```

```{r}
mob.lm3$coefficients
```

## Or even


```{r}
summary(mob.lm3)$coef
```

## Confidence Intervals

* If we want confidence intervals for the coefficients, we can use `confint`:

```{r}
confint(mob.lm3, level = 0.90) # default confidence level is 0.95
```

## Warning!!


* This calculates confidence intervals assuming independent, constant-variance
Gaussian noise everywhere, etc., etc., so it's not to be taken too seriously
unless you've checked those assumptions somehow; 


## Fitted values and residuals

For every data point in the original data set, we have both a fitted
value ($\widehat{y}$) and a residual ($y-\widehat{y}$).  These are vectors, and
can be extracted with the `fitted` and `residuals` functions:

```{r}
head(fitted(mob.lm2)) # head() gives the first few elements
tail(residuals(mob.lm2)) # tail() gives the last few elements
```


## Using bits of the lm output


* You may be more used to accessing all these things as parts of the estimated
model --- writing something like `mob.lm2$coefficients` to get the coefficients.

* This is fine as far as it goes, but we will work with many different sorts of
statistical models in this course, and those internal names can change from
model to model.  

* If the people implementing the models did their job, however,
functions like `fitted`, `residuals`, `coefficients` and `confint` will all, to
the extent they apply, work, and work in the same way.

```{r}
names(mob.lm2)
```


## Methods and Classes (R-Geeky But Important)


* In R things like `residuals` or `coefficients` are a special kind of function,
called **methods**. (in `python`, these are like `obj.residuals`. The dot after the object accesses methods defined for that type of object.)  

* Other methods, which you've used a lot without perhaps
realizing it, are `plot`, `print` and `summary`.  

* These are a sort of generic
or meta-function, which looks up the class of model being used, and then calls
a specialized function which how to work with that class.  

* The convention is
that the specialized function is named _method_`.`_class_, e.g., `summary.lm`.

* If no specialized function is defined, R will try to use _method_`.default`.

## Wherefore methods?


* The advantage of methods is that you, as a user, don't have to learn a totally
new syntax to get the coefficients or residuals of every new model class

* you just use `residuals(mdl)` whether `mdl` comes from a linear regression which
could have been done two centuries ago, or is a Batrachian Emphasis Machine
which won't be invented for another five years.  

* (It also means that core parts
of R don't have to be re-written every time someone comes up with a new model
class.)  

* The one draw-back is that the help pages for the generic methods tend
to be pretty vague, and you may have to look at the help for the class-specific
functions 

* Compare `?summary` with `?summary.lm`.  

(If you are not sure what
the class of your model, `mdl`, is called, use `class(mdl)`.)




## Making Predictions


* The point of a regression model is to do prediction, and the method for doing
so is, naturally enough, called `predict`.  It works like so:  
```{r, eval=FALSE}
predict(object, newdata)
```

* Here `object` is an already estimated model, and `newdata` is a data frame
containing the new cases, real or imaginary, for which we want to make
predictions.  

* The output is (generally) a vector, with a predicted value for
each row of `newdata`.  

* If the rows of `newdata` have names, those will be
carried along as names in the output vector.  

```{r}
predict(mob.lm2, newdata = mob[which(mob$State == "AL"),])
```

```{r, message=FALSE}
library(tidyverse)
predict(mob.lm2, newdata = filter(mob, State == "AL"))
```

## Remember


* It is important to remember that making a prediction does _not_ mean "changing
the data and re-estimating the model"; 

* It means taking the unchanged estimate
of the model, and putting in new values for the covariates or independent
variables.  

* (In terms of the linear model, we change $x$, not $\widehat{\beta}$.)

* Notice that I used `mob.lm2` here, rather than the mathematically-equivalent `mob.lm1`.

* Because I specified `mob.lm2` with a formula that just referred to column names,
`predict` looks up columns with those names in `newdata`, puts them into the
function estimated in `mob.lm2`, and calculates the predictions.  

* Had I tried to
use `mob.lm1`, it would have completely ignored `newdata`.  

* This is one crucial
reason why it is best to use clean formulas and a `data` argument when
estimating the model.

## Transformations


* If the formula specifies transformations, those will also be done on `newdata`;

* we don't have to do the transformations ourselves:

```{r}
predict(mob.lm3, newdata = filter(mob, State=="AL"))
```

* The `newdata` does not have to be a subset of the original data used for
estimation, or related to it in any way at all

## Fun with predict


* It just has to have columns
whose names match those in the right-hand side of the formula.

```{r}
predict(
  mob.lm3, 
  newdata = data.frame(
    Population=1.5e6, Seg_racial=0, Commute=0.5, 
    Income=3e4, Gini = median(mob$Gini)
  )
)
predict(
  mob.lm3, 
  newdata = data.frame(
    Population=1.5e6, Seg_racial=0, 
    Commute=0.5, Income = quantile(mob$Income, c(0.05,0.5,0.95)),
    Gini=quantile(mob$Gini,c(0.05,0.5,0.95))
  )
)
```

## Problems w/ predict


* A very common programming error is to run `predict` and get out a vector whose length equals the number of rows in the original estimation data

* and which
doesn't change no matter what you do to `newdata`.  

* This is because if `newdata` is missing, or if R cannot find all the variables it needs in it, the default is the predictions of the model on the original data.

* An even more annoying form of this error consists of forgetting that the
argument is called `newdata` and not `data`:

```{r}
head(predict(mob.lm3)) # Equivalent to head(fitted(mob.lm3))
```

## More problems


```{r}
head(predict(
  mob.lm3,
  data=data.frame(Population=1.5e6, Seg_racial=0, 
                  Commute=0.5, Income=3e4, Gini=median(mob$Gini))
  )
) 
# Don't do this!
```

* Returning the original fitted values when `newdata` is missing or messed up is
not what I would have chosen, but nobody asked me.

* Because `predict` is a method, the generic help file is fairly vague, and many
options are only discussed on the help pages for the class-specific functions

* compare `?predict` with `?predict.lm`.  

* Common options include giving standard errors for predictions (as well point forecasts), and giving various sorts of intervals.


## Using Different Model Classes


* All of this carries over to different model classes, at least if they've been well-designed.  

* For instance, suppose we want to estimate a kernel regression
on the same data, using the same variables.  

```{r test, message = FALSE, warning=FALSE, results="hide"}
library(np)
mob.npbw <- npregbw(formula=formula(mob.lm2), data=mob, tol=1e-2, ftol=1e-2)
mob.np <- npreg(mob.npbw, data=mob)
```

(no need to know what all these arguments mean at the moment.)


## Why this is easy

* We can re-use the formula, because it's just saying what the input and target variables of the regression are, and we want that to stay the same.  

* More importantly, both `lm` and `npreg` use the same mechanism, of separating the formula specifying the model from the data set containing the actual values of the variables.  

* Of course, some models have variations in allowable formulas  

    - interactions make sense for `lm` but not for `npreg`, 
    - the latter has a special way of dealing with ordered categorical variables that `lm` doesn't
    - etc.

* After estimating the model, we can do most of the same things to it that we could do to a linear model.  

## We can look at a summary:


```{r}
summary(mob.np)
```

## We can look at fitted values and residuals:

```{r}
head(fitted(mob.np))
tail(residuals(mob.np))
```

## We can make predictions:

```{r}
predict(mob.np, newdata=data.frame(Population=1.5e6, Seg_racial=0, 
        Commute=0.5, Income=3e4, Gini=median(mob$Gini)))
```

## and we can plot things

```{r, fig.align='center',fig.height=12,fig.width=10}
par(mar=c(5,5,1,1),cex.lab=3,cex.axis=2,lwd=2,col=4,bty='n')
plot(mob.np,plot.errors.method='bootstrap')
```