---
lecture: "00 Quiz 0 fun"
format: revealjs
metadata-files: 
  - _metadata.yml
---

{{< include _titleslide.qmd >}}


## Why this class?

- "This is a required course"
- "Heard it's fun!"
- "I want to graduate"
- "8am!" (/s)
- "I want to learn a lot about ML."
- "I want to dig deeper after taking 306."
- "I want to get... more comfortable with R."
- "At first it was because everyone I knew was taking it."
- "Practical skills in preperation for future careers"


## Syllabus Q

```{r load-quiz, include=FALSE}
library(tidyverse)
quiz <- tibble(read.csv("canvas_quiz0.csv"))
```

```{r}
#| echo: false
#| fig-height: 6
#| fig-width: 12
quiz |> 
  ggplot(aes(str_wrap(syllabus_fact, 20))) + 
  geom_bar(fill = "blue") + 
  scale_y_continuous(expand = expansion(c(0,0.05))) +
  xlab("") 
```


## Programming languages

```{r}
#| echo: false
#| fig-height: 6
#| fig-width: 12
quiz |> 
  ggplot(aes(str_wrap(r_experience, 20))) + 
  geom_bar(fill = "blue") + 
  scale_y_continuous(expand = expansion(c(0, 0.05))) +
  xlab("")
```

<!--
## Matrix inversion

```{r}
library(MASS)
X <- matrix(c(5, 3, 1, -1), nrow = 2)
X
solve(X)
ginv(X)
X^(-1)
```



## Linear models

```{r}
y <- X %*% c(2, -1) + rnorm(2)
coefficients(lm(y ~ X))
coef(lm(y ~ X))
solve(t(X) %*% X) %*% t(X) %*% y
solve(crossprod(X), crossprod(X, y))
```

```{r}
#| error: true
X \ y # this is Matlab
```
---
-->

## Plans

```{r}
#| echo: false
#| fig-height: 6
#| fig-width: 12
library(cowplot)

quiz %>%
  ggplot(aes(str_wrap(graduation_plans, 20))) +
  geom_bar(fill = "orange") +
  scale_y_continuous(expand = expansion(c(0, 0.05))) +
  xlab("")
```

## Grade predictions

```{r, echo=FALSE, fig.height=4, fig.width=12, fig.align='center'}
#| echo: false
#| fig-height: 4
#| fig-width: 12
quiz |>
  ggplot(aes(predicted_grade)) +
  geom_histogram(
    fill = "orange", colour = "black", 
    breaks = c(0, 50, 54, 59, 63, 67, 71, 75, 79, 84, 89, 101)
  ) +
  scale_y_continuous(expand = expansion(c(0, 0.05)))
```

* `r sum(quiz$predicted_grade == 100)` people say 100%
* `r sum(quiz$predicted_grade == 90)` say 90%
* `r sum(quiz$predicted_grade == 85)` say 85%
* `r sum(quiz$predicted_grade == 80)` say 80%
