---
lecture: "00 Quiz 0 fun"
format: revealjs
metadata-files: 
  - _metadata.yml
---

{{< include _titleslide.qmd >}}


## Why this class?

* Most say requirements.
* Interest in ML/Stat learning
* Expressions of love/affection for Stats/CS/ML
* Enjoyment of past similar classes

## Why this class?

More idiosyncratic:

::: incremental

- "Professor received Phd from CMU, must be an awesome  researcher."
- "Learn strategies."
- (paraphrase) "Course structure with less weight on exam helps with anxiety"
- (paraphrase) "I love coding in R and want more of it"
- "Emmmmmmmmmmmmmmmm, to learn some skills from Machine Learning and finish my minor🙃."
- "destiny"
- "challenges from ChatGPT"
- "I thought Daniel Mcdonald is a cool prof..."
- "I have heard this is the most useful stat course in UBC."

:::


## Syllabus Q

```{r load-quiz, include=FALSE}
quiz <- read_rds(here::here(
  "..", "..", "..", "WinterT1-2023", "gh-class-management", "lab0",
  "lab0-anon.rds"
))
```

```{r}
#| echo: false
#| fig-height: 6
#| fig-width: 12
quiz |> 
  ggplot(aes(str_wrap(syllabus, 20))) + 
  geom_bar(fill = blue) + 
  scale_y_continuous(expand = expansion(c(0,0.05))) +
  xlab("") 
```


## Programming languages

```{r}
#| echo: false
#| fig-height: 6
#| fig-width: 12
quiz |> 
  ggplot(aes(str_wrap(r_coding, 20))) + 
  geom_bar(fill = blue) + 
  scale_y_continuous(expand = expansion(c(0, 0.05))) +
  xlab("")
```

---

## Matrix inversion

```{r}
library(MASS)
X <- matrix(c(5, 3, 1, -1), nrow = 2)
X
solve(X)
ginv(X)
X^(-1)
```



## Linear models

```{r}
y <- X %*% c(2, -1) + rnorm(2)
coefficients(lm(y ~ X))
coef(lm(y ~ X))
solve(t(X) %*% X) %*% t(X) %*% y
solve(crossprod(X), crossprod(X, y))
```

```{r}
#| error: true
X \ y # this is Matlab
```


## Pets and plans

```{r}
#| echo: false
#| fig-height: 6
#| fig-width: 12
library(cowplot)
p1 <- quiz %>%
  ggplot(aes(str_wrap(pets, 30))) +
  geom_bar(fill = blue) +
  coord_flip() +
  scale_y_continuous(expand = expansion(c(0, 0.05))) +
  xlab("")
p2 <- quiz %>%
  ggplot(aes(str_wrap(plans, 20))) +
  geom_bar(fill = orange) +
  coord_flip() +
  scale_y_continuous(expand = expansion(c(0, 0.05))) +
  xlab("")
plot_grid(p1, p2)
```

---

## Grade predictions

```{r, echo=FALSE, fig.height=4, fig.width=12, fig.align='center'}
#| echo: false
#| fig-height: 4
#| fig-width: 12
quiz |>
  ggplot(aes(grade)) +
  geom_histogram(
    fill = orange, colour = "black", 
    breaks = c(0, 50, 54, 59, 63, 67, 71, 75, 79, 84, 89, 101)
  ) +
  scale_y_continuous(expand = expansion(c(0, 0.05)))
```

* `r sum(quiz$grade == 100)` people say 100%
* `r sum(quiz$grade == 90)` say 90%
* `r sum(quiz$grade == 85)` say 85%
* `r sum(quiz$grade == 80)` say 80%
* Lots of clumping


. . .

1 said 35, and 1 said 50. Woof!

## Prediction accuracy (last year)

```{r}
#| echo: false
acc <- read_rds(here::here(
  "..", "..", "..", "WinterT1-2022", "final-grades",
  "prediction-accuracy.rds"
)) |>
  rename(predicted = pfg, actual = fg)

ggplot(acc, aes(predicted, actual)) +
  geom_point(colour = orange) +
  geom_abline(slope = 1, intercept = 0) +
  coord_equal(xlim = c(0, 100), ylim = c(0, 100))
```

## Prediction accuracy (last year)

```{r}
summary(lm(actual ~ predicted - 1, data = acc))
```
