---
lecture: "05 Estimating test MSE"
format: revealjs
metadata-files: 
  - _metadata.yml
---

{{< include _titleslide.qmd >}}


## Estimating prediction risk

Last time, we saw

$R_n(\widehat{f}) = E[(Y-\widehat{f}(X))^2]$


prediction risk  =  $\textrm{bias}^2$  +  variance  +  irreducible error 


We argued that we want procedures that produce $\widehat{f}$ with small $R_n$.


> How do we estimate $R_n$?



## Don't use training error


The training error in regression is

$$\widehat{R}_n(\widehat{f}) = \frac{1}{n} \sum_{i=1}^n (y_i - \hat{f}(x_i))^2$$

Here, the $n$ is doubly used (annoying, but simple): $n$ observations to create $\widehat{f}$ and $n$ terms in the sum.

::: callout-important
Training error is a bad estimator for $R_n(\widehat{f})$.
:::


So we should [__never__]{.secondary} use it.

## These all have the same $R^2$ and Training Error


::: flex
::: w-50

```{r}
#| code-fold: true
#| fig-height: 8
#| fig-width: 8
ans <- anscombe |>
  pivot_longer(everything(), names_to = c(".value", "set"), 
               names_pattern = "(.)(.)")
ggplot(ans, aes(x, y)) + 
  geom_point(colour = orange, size = 3) + 
  geom_smooth(method = "lm", se = FALSE, color = blue, linewidth = 2) +
  facet_wrap(~set, labeller = label_both)
```

:::
::: w-50

```{r}
ans %>% 
  group_by(set) |> 
  summarise(
    R2 = summary(lm(y ~ x))$r.sq, 
    train_error = mean((y - predict(lm(y ~ x)))^2)
  ) |>
  kableExtra::kable(digits = 2)
```

:::
:::

## Adding "junk" predictors increases $R^2$ and decreases Training Error

```{r}
n <- 100
p <- 10
q <- 0:30
x <- matrix(rnorm(n * (p + max(q))), nrow = n)
y <- x[, 1:p] %*% c(5:1, 1:5) + rnorm(n, 0, 10)

regress_on_junk <- function(q) {
  x <- x[, 1:(p + q)]
  mod <- lm(y ~ x)
  tibble(R2 = summary(mod)$r.sq,  train_error = mean((y - predict(mod))^2))
}
```

```{r}
#| code-fold: true
map(q, regress_on_junk) |> 
  list_rbind() |>
  mutate(q = q) |>
  pivot_longer(-q) |>
  ggplot(aes(q, value, colour = name)) +
  geom_line(linewidth = 2) + xlab("train_error") +
  scale_colour_manual(values = c(blue, orange), guide = "none") +
  facet_wrap(~ name, scales = "free_y")
```


## Other things you can't use

You should not use `anova` 

or the $p$-values from the `lm` output for this purpose.

> These things are to determine whether those _parameters_ are different from zero if you were to repeat the experiment many times, if the model were true, etc. etc.

Not the same as _"are they useful for prediction = do they help me get smaller $R_n$?"_



## Risk of Risk

While it's crummy, Training Error is an [estimator]{.secondary} of $R_n(\hat{f})$

Recall, $R_n(\hat{f})$ is a _parameter_ (a property of the data distribution)

So we can ask "is $\widehat{R}(\hat{f})$ a good estimator for $R_n(\hat{f})$?"

Both are just numbers, so perhaps a good way to measure is

$$
E[(R_n - \widehat{R})^2] 
= \cdots 
= (R_n - E[\widehat{R}])^2 + \Var{\widehat{R}}
$$

Choices you make determine how good this is.

We can try to balance it's bias and variance...

# Things you can use.


## Held out sets

One option is to have a separate "held out" or "validation set".


👍 Estimates the test error

👍 Fast computationally

🤮 Estimate is random 

🤮 Estimate has high variance (depends on 1 choice of split)

🤮 Estimate has some bias because we only used some of the data


## Aside {background-color="#97D4E9"}

In my experience, CS has particular definitions of "training", "validation", and "test" data.

I think these are not quite the same as in Statistics.

* [Test data]{.secondary} - Hypothetical data you don't get to see, ever. Infinite amounts drawn from the population.
    * _Expected test error_ or _Risk_ is an expected value over this distribution. It's _not_ a sum over some data kept aside.
* Sometimes I'll give you "test data". You pretend that this is a good representation of the expectation and use it to see how well you did on the training data.
* [Training data]{.secondary} - This is data that you get to touch.
* [Validation set]{.secondary} - Often, we need to _choose models_. One way to do this is to split off some of your training data and pretend that it's like a "Test Set".

When and how you split your training data can be very important. 



## Intuition for CV


One reason that $\widehat{R}_n(\widehat{f})$ is bad is that we are using the same data to pick $\widehat{f}$ __AND__ to estimate $R_n$.

"Validation set" fixes this, but holds out a particular, fixed block of data we pretend mimics the "test data"

. . .

What if we set aside one observation, say the first one $(y_1, x_1)$.

We estimate $\widehat{f}^{(1)}$ without using the first observation.

Then we test our prediction:

$$\widetilde{R}_1(\widehat{f}^{(1)}) = (y_1 -\widehat{f}^{(1)}(x_1))^2.$$


(why the notation $\widetilde{R}_1$? Because we're estimating the risk with 1 observation. )

---

## Keep going

But that was only one data point $(y_1, x_1)$. Why stop there?

Do the same with $(y_2, x_2)$! Get an estimate $\widehat{f}^{(2)}$ 
without using it, then

$$\widetilde{R}_1(\widehat{f}^{(2)}) = (y_2 -\widehat{f}^{(2)}(x_2))^2.$$

We can keep doing this until we try it for every data point.

And then average them! (Averages are good)


$$\mbox{LOO-CV} = \frac{1}{n}\sum_{i=1}^n \widetilde{R}_1(\widehat{f}^{(i)}) = \frac{1}{n}\sum_{i=1}^n 
(y_i - \widehat{f}^{(i)}(x_i))^2$$

. . .

This is [__leave-one-out cross validation__]{.secondary}


## Problems with LOO-CV

🤮 Each held out set is small $(n=1)$. Therefore, the variance of the Squared Error of each prediction is high.

🤮 The training sets overlap. This is bad. 

- Usually, averaging reduces variance: $\Var{\overline{X}} = \frac{1}{n^2}\sum_{i=1}^n \Var{X_i} = \frac{1}{n}\Var{X_1}.$
- But only if the variables are independent. If not, then $\Var{\overline{X}} = \frac{1}{n^2}\Var{ \sum_{i=1}^n X_i} = \frac{1}{n}\Var{X_1} + \frac{1}{n^2}\sum_{i\neq j} \Cov{X_i}{X_j}.$
- Since the training sets overlap a lot, that covariance can be pretty big.
    
🤮 We have to estimate this model $n$ times.

🎉 Bias is low because we used almost all the data to fit the model: $E[\mbox{LOO-CV}] = R_{n-1}$ 

  
## K-fold CV

::: flex
::: w-50
To alleviate some of these problems, people usually use $K$-fold cross validation.

The idea of $K$-fold is 

1. Divide the data into $K$ groups. 
1. Leave a group out and estimate with the rest.
1. Test on the held-out group. Calculate an average risk over these $\sim n/K$ data.
1. Repeat for all $K$ groups.
1. Average the average risks.


:::


::: w-50
🎉 Less overlap, smaller covariance.

🎉 Larger hold-out sets, smaller variance.

🎉 Less computations (only need to estimate $K$ times)

🤮 LOO-CV is (nearly) unbiased for $R_n$

🤮 K-fold CV is unbiased for $R_{n(1-1/K)}$

The risk depends on how much data you use to estimate the model. $R_n$ depends on $n$.

:::
:::



## A picture

```{r}
#| code-fold: true
#| fig-height: 6
#| fig-width: 10
par(mar = c(0, 0, 0, 0))
plot(NA, NA, ylim = c(0, 5), xlim = c(0, 10), bty = "n", yaxt = "n", xaxt = "n")
rect(0, .1 + c(0, 2, 3, 4), 10, .9 + c(0, 2, 3, 4), col = blue, density = 10)
rect(c(0, 1, 2, 9), rev(.1 + c(0, 2, 3, 4)), c(1, 2, 3, 10), 
     rev(.9 + c(0, 2, 3, 4)), col = red, density = 10)
points(c(5, 5, 5), 1 + 1:3 / 4, pch = 19)
text(.5 + c(0, 1, 2, 9), .5 + c(4, 3, 2, 0), c("1", "2", "3", "K"), cex = 3, 
     col = red)
text(6, 4.5, "Training data", cex = 3, col = blue)
text(2, 1.5, "Validation data", cex = 3, col = red)
```


## Code


```{r}
#| code-line-numbers: "11-13|14-16|"
#' @param data The full data set
#' @param estimator Function. Has 1 argument (some data) and fits a model. 
#' @param predictor Function. Has 2 args (the fitted model, the_newdata) and produces predictions
#' @param error_fun Function. Has one arg: the test data, with fits added.
#' @param kfolds Integer. The number of folds.
kfold_cv <- function(data, estimator, predictor, error_fun, kfolds = 5) {
  n <- nrow(data)
  fold_labels <- sample(rep(1:kfolds, length.out = n))
  errors <- double(kfolds)
  for (fold in seq_len(kfolds)) {
    test_rows <- fold_labels == fold
    train <- data[!test_rows, ]
    test <- data[test_rows, ]
    current_model <- estimator(train)
    test$.preds <- predictor(current_model, test)
    errors[fold] <- error_fun(test)
  }
  mean(errors)
}
```

. . . 

```{r}
#| code-line-numbers: "2-4|"
somedata <- data.frame(z = rnorm(100), x1 = rnorm(100), x2 = rnorm(100))
est <- function(dataset) lm(z ~ ., data = dataset)
pred <- function(mod, dataset) predict(mod, newdata = dataset)
error_fun <- function(testdata) mutate(testdata, errs = (z - .preds)^2) |> pull(errs) |> mean()
kfold_cv(somedata, est, pred, error_fun, 5)
```


## Trick

__For a certain "nice" models__, one can show 

(after pages of tedious algebra which I wouldn't wish on my worst enemy, but might, in a fit of rage assign as homework to belligerent students) 


$$\mbox{LOO-CV} = \frac{1}{n} \sum_{i=1}^n \frac{(y_i -\widehat{y}_i)^2}{(1-h_{ii})^2} = \frac{1}{n} \sum_{i=1}^n \frac{\widehat{e}_i^2}{(1-h_{ii})^2}.$$

* This trick means that you only have to fit the model once rather than $n$ times!

* You still have to calculate this for each model!

```{r}
cv_nice <- function(mdl) mean( (residuals(mdl) / (1 - hatvalues(mdl)))^2 )
```

## Trick

```{r}
cv_nice <- function(mdl) mean( (residuals(mdl) / (1 - hatvalues(mdl)))^2 )
```


"Nice" requires:

1. $\widehat{y}_i = h_i(\mathbf{X})^\top \mathbf{y}$ for some vector $h_i$
2. $e^{(i)} = \frac{\widehat{e}_i}{(1-h_{ii})}$


# Next time...

More tricks and what's up with the name "hatvalues"
