---
title: "06 Information criteria"
author: 
  - "STAT 406"
  - "Daniel J. McDonald"
date: 'Last modified - `r Sys.Date()`'
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
source("rmd_config.R")
```

```{r css-extras, file="css-extras.R", echo=FALSE}
```


## Generalized CV

Last time we saw a nice trick

If $\widehat{f}(x_i) = \widehat{y}_i = h_i^\top \mathbf{y}$, for some vector $h_i(\mathbf{X})$

--

.hand[Same as "if] $\widehat{f}(\mathbf{X}) = \mathbf{H}\mathbf{y}$ .hand[for some matrix] $\mathbf{H}$ .hand["]


$$\mbox{LOO-CV} = \frac{1}{n} \sum_{i=1}^n \frac{(y_i -\widehat{y}_i)^2}{(1-h_{ii})^2} = \frac{1}{n} \sum_{i=1}^n \frac{\widehat{e}_i^2}{(1-h_{ii})^2}.$$

--

### This is another nice trick.

Idea: replace $h_{ii}$ with $\frac{1}{n}\sum_{i=1}^n h_{ii} = \frac{1}{n}\textrm{tr}(\mathbf{H})$

Let's call $\textrm{tr}(\mathbf{H})$ the __effective degrees-of-freedom__ (or just __df__)

$$\textrm{GCV} = \frac{1}{n} \sum_{i=1}^n \frac{\widehat{e}_i^2}{(1-\textrm{df}/n)^2} = \frac{\textrm{MSE}}{(1-\textrm{df}/n)^2}$$

--

.hand[Where does this stuff come from?]

$$
\newcommand{\Expect}[1]{E\left[ #1 \right]}
\newcommand{\Var}[1]{\mathrm{Var}\left[ #1 \right]}
\newcommand{\Cov}[2]{\mathrm{Cov}\left[#1,\ #2\right]}
\newcommand{\given}{\ \vert\ }
\newcommand{\y}{\mathbf{y}}
\newcommand{\X}{\mathbf{X}}
\newcommand{\H}{\mathbf{H}}
$$

---

## What are `hatvalues`?

```{r}
cv_nice <- function(mdl) mean((residuals(mdl) / (1 - hatvalues(mdl)))^2)
```

In OLS, $\widehat{\y} = \X\widehat{\beta} = \X(\X^\top \X)^{-1}\X^\top \y$

We often call $\H = \X(\X^\top \X)^{-1}\X^\top$ the Hat matrix, because it .hand[puts the hat] on $\y$

GCV uses $\textrm{tr}(\H)$. For `lm`, this is just `p`, the number of predictors (Why?)

This is one way of understanding the name __effective degrees-of-freedom__

---

## Alternative interpretation:

Suppose, $Y_i$ comes (independently) from some distribution and has mean $\mu_i$ and variance $\sigma^2$

--

(remember: in the linear model $\Expect{Y_i} = x_i^\top \beta = \mu_i$ )

--

Let $\widehat{Y}$ be an estimator of $\mu$ (all $i=1,\ldots,n$ elements of the vector).

\begin{aligned}
\Expect{\frac{1}{n}\sum (\widehat Y_i-\mu_i)^2} 
&= \Expect{\frac{1}{n}\sum (\widehat Y_i-Y_i + Y_i -\mu_i)^2}\\
&= \frac{1}{n}\Expect{\sum (\widehat Y_i-Y_i)^2} + \frac{1}{n}\Expect{\sum (Y_i-\mu_i)^2} + \frac{2}{n}\Expect{\sum (\widehat Y_i-Y_i)(Y_i-\mu_i)}\\
&= \frac{1}{n}\sum \Expect{(\widehat Y_i-Y_i)^2} + \sigma^2 + \frac{2}{n}\Expect{\sum (\widehat Y_i-Y_i)(Y_i-\mu_i)}\\
&\quad\vdots \\
&= \frac{1}{n}\sum \Expect{(\widehat Y_i-Y_i)^2} - \sigma^2 + \frac{2}{n}\sum\Cov{Y_i}{\widehat Y_i}
\end{aligned}

--

Now, if $\widehat{Y} = \H Y$ for some matrix $\H$,

$\sum\Cov{Y_i}{\widehat Y_i} = \Expect{Y^\top \H Y} = \sigma^2 \textrm{tr}(\H)$

--

This gives __Mallow's $C_p$__ aka __Stein's Unbiased Risk Estimator__:

$MSE + 2\hat{\sigma}^2\textrm{df}/n$

---

## AIC and BIC

These have a very similar flavor to $C_p$, but their genesis is different.

Without going into too much detail, they look like

$\textrm{AIC}/n = -2\textrm{loglikelihood}/n + 2\textrm{df}/n$

$\textrm{BIC}/n = -2\textrm{loglikelihood}/n + 2\log(n)\textrm{df}/n$

--

In the case of a linear model with Gaussian errors, 

$\textrm{AIC} = -n + 2n\log(2\pi) - 2 + 2\log(n) - 2\log(RSS) + 2(p+2) \propto -2\log(RSS) + 2(p+2)$

( $p+2$ because of the intercept and the unknown variance)
--

.alert[
Unfortunately, different books/software/notes define these differently. Even different R packages. This is __super annoying__. 

Forms above are in [ESL] eq. (7.29) and (7.35). [ISLR] gives special cases in Section 6.1.3. Remember the generic form here.
]

---

## Over-fitting vs. Under-fitting

.emphasis[
Over-fitting means estimating a really complicated function when you don't have enough data.
]

This is likely a .hand[low-bias / high-variance] situation.

.emphasis[
Under-fitting means estimating a really simple function when you have lots of data. 
]

This is likely a .hand[high-bias / low-variance] situation.

Both of these outcomes are bad (they have high risk $=$ big $R_n$ ).

The best way to avoid them is to use a reasonable estimate of __prediction risk__ to choose how complicated your model should be.

---

## Recommendations

.emphasis[
When comparing models, choose one criterion: CV / AIC / BIC / Cp / GCV. 

CV is usually easiest to make sense of and doesn't depend on other unknown parameters.

But, it requires refitting the model.

Also, it can be strange in cases with discrete predictors, time series, repeated measurements, graph structures, etc.
]

--

High-level intuition of these:

* GCV tends to choose "dense" models.

* Theory says AIC chooses the "best predictor" asymptotically.

* Theory says BIC should choose the "true model" asymptotically, tends to select fewer predictors.

* In some special cases, AIC = Cp = SURE $\approx$ LOO-CV

--

For more information: see [ESL] Chapter 7. And also <https://arxiv.org/abs/1602.01522>.

This material is more challenging than the level of this course, and is easily and often misunderstood.

--

.hand-blue[My recommendation: Use CV]

---

## A few more caveats

It is often tempting to "just compare" risk estimates from vastly different models. 

For example, 

* different transformations of the predictors, 

* different transformations of the response, 

* Poisson likelihood vs. Gaussian likelihood in `glm()`

.emphasis[
**This is not always justified.** 

1. The "high-level intuition" is for "nested" models.

1. Different likelihoods aren't comparable.

1. Residuals / response variables on different scales aren't directly comparable.
]

"Validation set" is easy, because you're always comparing to the "right" thing. But it has lots of drawbacks.
---
class: middle, inverse, center

# Next time ...

Greedy selection