---
lecture: "06 Information Criteria"
format: revealjs
metadata-files: 
  - _metadata.yml
---

{{< include _titleslide.qmd >}}

## LOO-CV

- Train $\hat f_1$ on all but data point 1, calculate $\tilde R_1 = \ell(Y_1, \hat f_1(X_1))$.
- Do the same for each data point $i$, calculate $\tilde R_i$
- Estimate $R_n \approx \hat R_n = \frac{1}{n}\sum_{i=1}^n \tilde R_i$

Has low bias 🎉 and (probably) low variance 🎉 

**Not often possible to use**: requires training $n$ models 🤮

## LOO-CV: Math to the rescue!

Consider models where predictions are a **linear function** of the training responses, i.e.,

$$ \hat{\mathbf y} = {\mathbf H} {\mathbf y} $$

where we collected terms into matrices and vectors (${\mathbf h_i}$ can be any functions):

- $\hat{\mathbf y} = \begin{bmatrix} \hat Y_1 & \cdots & \hat Y_n \end{bmatrix}^\top \in \mathbb R^{n}$
- ${\mathbf y} = \begin{bmatrix} Y_1 & \cdots & Y_n \end{bmatrix}^\top \in \mathbb R^{n}$
- $\mathbf H = \begin{bmatrix} \mathbf h_1(X_{1:n}) & \cdots & \mathbf h_n(X_{1:n}) \end{bmatrix}^\top \in \mathbb R^{n \times n}$

. . .

For example, OLS:

$$ \hat{\mathbf y} = {\mathbf X} \hat \beta, \qquad \hat\beta = (\mathbf X^\top \mathbf X)^{-1} \mathbf X^\top \mathbf y $$

By inspection $\mathbf H = \mathbf X (\mathbf X^\top \mathbf X)^{-1} \mathbf X^\top$


## LOO-CV: Math to the rescue!

For models where predictions are a **linear function** of the training responses\*,

**LOO-CV has a closed-form expression!** Just need to fit *once*:

$$\mbox{LOO-CV} \,\, \hat R_n = \frac{1}{n} \sum_{i=1}^n \frac{(Y_i -\widehat{Y}_i)^2}{(1-{\boldsymbol H}_{ii})^2}.$$

- Numerator is the _squared residual_ (loss) for training point $i$.
- Denominator weights each residual by *diagonal of $H$* some factor 
- $H_{ii}$ are *leverage/hat values*: tell you what happens when moving data point $i$ a bit

\*: plus some technicalities 

. . .

:::callout-tip
Deriving this sucks. I wouldn't recommend doing it yourself. 
:::

## Computing the formula

```{r}
cv_nice <- function(mdl) mean((residuals(mdl) / (1 - hatvalues(mdl)))^2)
```

## What happens when we can't use the formula?

(And can we get a better intuition about what's going on?)

$$
\hat{\mathbf y} = \mathbf H \mathbf y,
\qquad
\mbox{LOO-CV} = \frac{1}{n} \sum_{i=1}^n \frac{(Y_i -\widehat{Y}_i)^2}{(1-{\boldsymbol H}_{ii})^2}
$$

Let's look at OLS again...
$$ \hat Y = X \hat \beta, \qquad \beta = (\mathbf X^\top \mathbf X)^{-1} \mathbf X^\top \mathbf y $$

This implies that $\mathbf H = \mathbf X (\mathbf X^\top \mathbf X)^{-1} \mathbf X^\top$

- One really nice property of $\mathbf H$ is that $\tr{\mathbf H} = p$ (where $X_n \in \R^p$). [(Why?)]{.secondary}

## Generalizing the LOO-CV formula

Let's call $\tr{\mathbf H} = p$ the _degrees-of-freedom_ (or just _df_) of our OLS estimator.\
[(Intuition: we have $p$ parameters to fit, or $p$ "degrees of freedom")]{.secondary}

\
**Idea:** in our LOO-CV formula, approximate each ${\mathbf H}_{ii}$ with the average $\frac 1 n \sum_{i=1}^n {\mathbf H}_{ii}$.\
\
Then...

$$
\mbox{LOO-CV} = \frac{1}{n} \sum_{i=1}^n \frac{(y_i -\widehat{y}_i)^2}{(1-{\mathbf H}_{ii})^2} \approx \frac{\text{MSE}}{(1-\text{df}/n)^2} \triangleq \text{GCV}
$$

GCV stands for **Generalized CV** estimate

<!--
## Generalized CV

Last time we saw a nice trick, that works some of the time (OLS, Ridge regression,...)

$$\mbox{LOO-CV} = \frac{1}{n} \sum_{i=1}^n \frac{(y_i -\widehat{y}_i)^2}{(1-h_{ii})^2} = \frac{1}{n} \sum_{i=1}^n \frac{\widehat{e}_i^2}{(1-h_{ii})^2}.$$

1. $\widehat{\y} = \widehat{f}(\mathbf{X}) = \mathbf{H}\mathbf{y}$ for some matrix $\mathbf{H}$.
2. A technical thing.

$$\newcommand{\H}{\mathbf{H}}$$

## This is another nice trick.

Idea: replace $h_{ii}$ with $\frac{1}{n}\sum_{i=1}^n h_{ii} = \frac{1}{n}\textrm{tr}(\mathbf{H})$

Let's call $\textrm{tr}(\mathbf{H})$ the _degrees-of-freedom_ (or just _df_)

$$\textrm{GCV} = \frac{1}{n} \sum_{i=1}^n \frac{\widehat{e}_i^2}{(1-\textrm{df}/n)^2} = \frac{\textrm{MSE}}{(1-\textrm{df}/n)^2}$$


[Where does this stuff come from?]{.hand}


## What are `hatvalues`?

```{r}
cv_nice <- function(mdl) mean((residuals(mdl) / (1 - hatvalues(mdl)))^2)
```

In OLS, $\widehat{\y} = \X\widehat{\beta} = \X(\X^\top \X)^{-1}\X^\top \y$

We often call $\mathbf{H} = \X(\X^\top \X)^{-1}\X^\top$ the Hat matrix, because it [puts the hat]{.hand} on $\y$

GCV uses $\textrm{tr}(\mathbf{H})$. 

For `lm()`, this is just `p`, the number of predictors (Why?)

This is one way of understanding the name _degrees-of-freedom_
-->

## Generalized CV

$$\textrm{GCV} = \frac{\textrm{MSE}}{(1-\textrm{df}/n)^2}$$

We can use this formula for models that aren't of the form $\widehat{y}_i = \boldsymbol h_i(\mathbf{X})^\top \mathbf{y}$.\
(Assuming we have some model-specific formula for estimating $\textrm{df}$.)

. . .

### Observations

- GCV > training error (Why?)
- What happens as $n$ increases?
- What happens as $\text{df}$ ($p$ in our OLS model) increases?


## Mallows $C_p$

Let's see if we can generalize risk estimators from OLS ($Y \sim \mathcal{N}(X^T\beta, \sigma^2)$) in other ways. 

Consider the *estimation risk* of estimating $\mu_i = X_i^T\beta$ with $\hat Y_i = X_i^T\hat\beta$: 

$$R_n = E\left[\frac{1}{n}\sum_{i=1}^n (\hat Y_i - \mu_i)^2\right]$$

Using the usual decomposition tricks:

$$
R_n= \Expect{\frac{1}{n}\sum (\widehat Y_i-\mu_i)^2} 
= \underbrace{\frac{1}{n}\sum \Expect{(\widehat Y_i-Y_i)^2}}_{\text{train MSE}} -
\underbrace{\sigma^2}_{\text{noise}} +
\underbrace{\frac{2}{n}\sum\Cov{Y_i}{\widehat Y_i}}_{\text{???}}
$$

<!--

\begin{aligned}
&= \Expect{\frac{1}{n}\sum (\widehat Y_i-Y_i + Y_i -\mu_i)^2}\\
&= \frac{1}{n}\Expect{\sum (\widehat Y_i-Y_i)^2} + \frac{1}{n}\Expect{\sum (Y_i-\mu_i)^2} + \frac{2}{n}\Expect{\sum (\widehat Y_i-Y_i)(Y_i-\mu_i)}\\
&= \frac{1}{n}\sum \Expect{(\widehat Y_i-Y_i)^2} + \sigma^2 + \frac{2}{n}\Expect{\sum (\widehat Y_i-Y_i)(Y_i-\mu_i)} = \cdots =\\
&= \frac{1}{n}\sum \Expect{(\widehat Y_i-Y_i)^2} - \sigma^2 + \frac{2}{n}\sum\Cov{Y_i}{\widehat Y_i}
\end{aligned}

## Alternative interpretation:

$$\Expect{\frac{1}{n}\sum (\widehat Y_i-\mu_i)^2} =
\underbrace{\frac{1}{n}\sum \Expect{(\widehat Y_i-Y_i)^2}}_{(1)} -
\underbrace{\sigma^2}_{(2)} +
\underbrace{\frac{2}{n}\sum\Cov{Y_i}{\widehat Y_i}}_{(3)}
$$

. . .

-->

## Mallows $C_p$

$$\Expect{\frac{1}{n}\sum (\widehat Y_i-\mu_i)^2} =
\underbrace{\frac{1}{n}\sum \Expect{(\widehat Y_i-Y_i)^2}}_{\text{training error}} -
\underbrace{\sigma^2}_{\text{noise}} +
\underbrace{\frac{2}{n}\sum\Cov{Y_i}{\widehat Y_i}}_{\text{???}}
$$

Recall that $\widehat{\mathbf{Y}} = \mathbf H \mathbf{Y}$ for some matrix $\mathbf H$,

$\sum\Cov{Y_i}{\widehat Y_i} = \Expect{\mathbf{Y}^\top \mathbf H \mathbf{Y}} = \sigma^2 \textrm{tr}(\mathbf H)$


This gives _Mallow's $C_p$_ aka _Stein's Unbiased Risk Estimator_:

$$ C_p = \text{MSE} + 2\hat{\sigma}^2 \: \textrm{df}/n $$

## Mallow's $C_p$

$$ C_p = \text{MSE} + 2\hat{\sigma}^2 \: \textrm{df}/n$$
(We derived it for the OLS model, but again it can be generalized to other models.)

::: callout-important
Unfortunately, $\text{df}$ may be difficult or impossible to calculate for complicated
prediction methods. But one can often estimate it well. This idea is beyond
the level of this course.
:::

### Observations
- $C_p$ > training error
- What happens as $n$ increases?
- What happens as $\text{df}$ ($p$ in our OLS model) increases?
- What happens as the irreducible noise increase?







# Last Day of Module 1
(Risk, CV, Information Criteria, Model/Variable Selection)

# HW 1 Due tonight (Sept 26) at 11pm

## Where We've Been (and What We Have Left)

::: flex
::: w-50

::: fragment
1. Risk (and test error) of a predictor:
    - $R_n = E[T_n]$
    - $R_n = \text{Est. Risk + Irr. Noise}$
    - $\text{Est. Risk} = \text{Bias}^2 + \text{Var}$
    - Best model = lowest risk
:::

::: fragment
2. Estimating risk with *holdout data*
    - ~~Estimate risk with training error~~ (bad!)
    - $k$-fold cross-validation (CV)
    - Leave-one-out CV (LOO-CV)
:::
:::

::: w-50

::: fragment
3. Estimating risk with *information criteria*
    - Information criteria
    - Mallows $C_p$
    - AIC/BIC
:::
  
::: fragment
4. Using risk estimates for *model/variable selection*
    - What is model selection?
    - Practical algorithms for model/variable selection
:::
  
:::
:::
  
## Cross Validation

```{r}
#| code-fold: true
#| fig-height: 6
#| fig-width: 10
par(mar = c(0, 0, 0, 0))
plot(NA, NA, ylim = c(0, 5), xlim = c(0, 10), bty = "n", yaxt = "n", xaxt = "n")
rect(0, .1 + c(0, 2, 3, 4), 10, .9 + c(0, 2, 3, 4), col = blue, density = 10)
rect(c(0, 1, 2, 9), rev(.1 + c(0, 2, 3, 4)), c(1, 2, 3, 10), 
     rev(.9 + c(0, 2, 3, 4)), col = red, density = 10)
points(c(5, 5, 5), 1 + 1:3 / 4, pch = 19)
text(.5 + c(0, 1, 2, 9), .5 + c(4, 3, 2, 0), c("1", "2", "3", "K"), cex = 3, 
     col = red)
text(6, 4.5, "Training data", cex = 3, col = blue)
text(2, 1.5, "Validation data", cex = 3, col = red)
```

- Unbiased estimate of $R_{n(1-1/K)}$ (but biased estimate of $R_n$)
- Requires training $K$ models

## GCV and Mallow's $C_p$

$$\textrm{GCV} = \frac{\textrm{MSE}}{(1-\textrm{df}/n)^2}$$
[(a hacked version of the magic LOO-CV formula)]{.small}

$$C_p = \mathrm{MSE} + 2 \hat{\sigma}^2 \textrm{df}/n$$
[(closed-form derivation of risk for OLS model with fixed $X$s)]{.small}

\

- $\mathrm{MSE} = \frac 1 N \sum_{i=1}^N (y_i - \hat f(x_i))^2$ is the *training error*
- $\mathrm{df}$ is the *degrees of freedom*
- For OLS, $\mathrm{df} = p$ (number of predictors)

## Akaike Information Criterion and Bayesian Information Criterion

These have a very similar flavor to $C_p$, but their genesis is different.

Without going into too much detail, they look like \


$\textrm{AIC}/n = -2\textrm{log-likelihood}/n + 2\textrm{df}/n$

$\textrm{BIC}/n = -2\textrm{log-likelihood}/n + 2\log(n)\textrm{df}/n$

[What is log-likelihood?]{.secondary}

. . .

In the case of $Y = X \hat \beta + \epsilon$ with $\epsilon \sim \mathcal{N}(0, \hat \sigma^2)$...


## AIC and BIC

In the case of $Y = X \hat \beta + \epsilon$ with $\epsilon \sim \mathcal{N}(0, \hat \sigma^2)$

$$ \text{log-likelihood} = -\frac{1}{2} \left( n\log(2\pi) + n\log(\hat \sigma^2) + \left(\textstyle \sum_{i=1}^n (Y_i - X\hat\beta)^2\right)/ \hat \sigma^2 \right) $$

- [$\sum_{i=1}^n (Y_i - X\hat\beta)^2 = n(\textrm{MSE})$]{.small}
- [$\hat \sigma^2 = \left(\textstyle \sum_{i=1}^n (Y_i - X\hat\beta)^2\right) / n = \textrm{MSE}$]{.small}

So, after simplifying...

$$\textrm{AIC}/n = \log(\textrm{MSE}) + 2 \textrm{df}/n + \mathrm{Const.}$$

$$\textrm{BIC}/n = \log(\textrm{MSE}) + 2 \log(n) \textrm{df}/n + \mathrm{Const.}$$

::: callout-important
Unfortunately, different books/software/notes define these differently. Even different R packages. This is __super annoying__. 
:::


## Summary

We have seen 5 risk estimators:

0. ~~Training error, R^2, p-values~~
1. k-fold CV / LOO-CV
2. Generalized CV
3. Mallows $C_p$
4. Akaike Information Criterion (AIC)
5. Bayesian Information Criterion (BIC)

. . .

### Geoff & Trevor's recommendation: 

**Use CV.** (LOO-CV if you have a magic formula, k-fold otherwise.)

[(So why did we just suffer through learning all those boring information criteria????)]{.secondary}


## Information criteria help us reason about models statistically

For regression models (with Gaussian noise), our information criteria are...
$$
\begin{align}
\textrm{GCV} &= \textrm{MSE} / (1-\textrm{df}/n)^2 \\
C_p &= \mathrm{MSE} + 2 \hat \sigma^2 \textrm{df}/n \\
\textrm{AIC}/n &= \log(\textrm{MSE}) + 2 \textrm{df}/n + \mathrm{Const.} \\
\textrm{BIC}/n &= \log(\textrm{MSE}) + 2 \log(n) \textrm{df}/n + \mathrm{Const.}
\end{align}
$$

### What is similar about these criteria?

### What is different?


## Over-fitting vs. Under-fitting

All 4 information criteria become large (i.e. we have high risk) when either

1. Training MSE is large
2. The ratio of $\textrm{df}/n$ is large

. . .

### The two failure modes of models

- **Over-fitting** means estimating a really complicated function when you don't have enough data.
- **Under-fitting** means estimating a really simple function when you have lots of data.

[For an overfit model, is MSE large or is $\textrm{df}/n$ large? What about for an underfit model?]{.secondary}

[Are overfit models high-bias or high-variance? What about underfit models?]{.secondary}


## Over-fitting vs. Under-fitting

- Overfitting = high variance = large $\textrm{df}/n$
- Underfitting = high biase = large training $\textrm{MSE}$

::: callout-tip
We have a diagonstic for determining if we're high bias versus high variance!\
Is training MSE large, or is $\textrm{df}/n$?
:::

The best way to avoid overfitting/underfitting is to use a reasonable estimate
of _prediction risk_ to choose how complicated your model should be.


<!--
## Commentary

- When comparing models, choose one criterion: CV / AIC / BIC / Cp / GCV. 
    - In some special cases, AIC = Cp = SURE $\approx$ LOO-CV
- CV is generic, easy, and doesn't depend on unknowns.
    - But requires refitting, and nontrivial for discrete predictors, time series, etc.
- GCV tends to choose "dense" models.
- Theory says AIC chooses "best predicting model" asymptotically.
- Theory says BIC chooses "true model" asymptotically, tends to select fewer predictors.
- Technical: CV (or validation set) is estimating error on 
[new data]{.secondary}, unseen $(X_0, Y_0)$; AIC / CP are estimating error on [new Y]{.secondary} at the observed $x_1,\ldots,x_n$. This is subtle.

::: aside
For more information: see [ESL] Chapter 7.
This material is more challenging than the level of this course, and is easily and often misunderstood.
:::



# My recommendation: 

**Use CV.**
-->

## A few more caveats

Tempting to "just compare" risk estimates from vastly different models. 

For example, 

* different transformations of the predictors, 

* different transformations of the response, 

* Poisson likelihood vs. Gaussian likelihood in `glm()`


[This is not always justified.]{.secondary}

1. We can compare "nested" models (i.e. where one model is a special case of another)

1. Different likelihoods aren't comparable.

1. Residuals / response variables on different scales aren't directly comparable.



# Next time ...

Using risk estimates for model selection.
