---
lecture: "10 Basis expansions"
format: revealjs
metadata-files: 
  - _metadata.yml
---

{{< include _titleslide.qmd >}}



## What about nonlinear things


$$\Expect{Y \given X=x} = \sum_{j=1}^p x_j\beta_j$$

Now we relax this assumption of linearity:

$$\Expect{Y \given X=x} = f(x)$$

How do we estimate $f$?

. . . 

For this lecture, we use $x \in \R$ (1 dimensional)

Higher dimensions are possible, but complexity grows [exponentially]{.secondary}.

We'll see some special techniques for $x\in\R^p$ later this Module.


## Start simple

For any $f : \R \rightarrow [0,1]$

$$f(x) = f(x_0) + f'(x_0)(x-x_0) + \frac{1}{2}f''(x_0)(x-x_0)^2 + \frac{1}{3!}f'''(x_0)(x-x_0)^3 + R_3(x-x_0)$$

So we can linearly regress $y_i = f(x_i)$ on the polynomials.

The more terms we use, the smaller $R$.

```{r}
#| code-fold: true
set.seed(406406)
data(arcuate, package = "Stat406") 
arcuate <- arcuate |> slice_sample(n = 220)
arcuate %>% 
  ggplot(aes(position, fa)) + 
  geom_point(color = blue) +
  geom_smooth(color = orange, formula = y ~ poly(x, 3), method = "lm", se = FALSE)
```


## Same thing, different orders

```{r}
#| code-fold: true
arcuate %>% 
  ggplot(aes(position, fa)) + 
  geom_point(color = blue) + 
  geom_smooth(aes(color = "a"), formula = y ~ poly(x, 4), method = "lm", se = FALSE) +
  geom_smooth(aes(color = "b"), formula = y ~ poly(x, 7), method = "lm", se = FALSE) +
  geom_smooth(aes(color = "c"), formula = y ~ poly(x, 25), method = "lm", se = FALSE) +
  scale_color_manual(name = "Taylor order",
    values = c(green, red, orange), labels = c("4 terms", "7 terms", "25 terms"))
```


## Still a "linear smoother"

Really, this is still linear regression, just in a transformed space.

It's not linear in $x$, but it is linear in $(x,x^2,x^3)$ (for the 3rd-order case)

So, we're still doing OLS with

$$\X=\begin{bmatrix}1& x_1 & x_1^2 & x_1^3 \\ \vdots&&&\vdots\\1& x_n & x_n^2 & x_n^3\end{bmatrix}$$

So we can still use our nice formulas for LOO-CV, GCV, Cp, AIC, etc.


```{r}
max_deg <- 20
cv_nice <- function(mdl) mean( residuals(mdl)^2 / (1 - hatvalues(mdl))^2 ) 
cvscores <- map_dbl(seq_len(max_deg), ~ cv_nice(lm(fa ~ poly(position, .), data = arcuate)))
```

## 

```{r}
#| code-fold: true
library(cowplot)
g1 <- ggplot(tibble(cvscores, degrees = seq(max_deg)), aes(degrees, cvscores)) +
  geom_point(colour = blue) +
  geom_line(colour = blue) + 
  labs(ylab = 'LOO-CV', xlab = 'polynomial degree') +
  geom_vline(xintercept = which.min(cvscores), linetype = "dotted") 
g2 <- ggplot(arcuate, aes(position, fa)) + 
  geom_point(colour = blue) + 
  geom_smooth(
    colour = orange, 
    formula = y ~ poly(x, which.min(cvscores)), 
    method = "lm", 
    se = FALSE
  )
plot_grid(g1, g2, ncol = 2)
```


## Other bases

Polynomials
: $x \mapsto \left(1,\ x,\ x^2, \ldots, x^p\right)$ (technically, not quite this, they are orthogonalized)

Linear splines
: $x \mapsto \bigg(1,\ x,\ (x-k_1)_+,\ (x-k_2)_+,\ldots, (x-k_p)_+\bigg)$ for some choices $\{k_1,\ldots,k_p\}$

Cubic splines
: $x \mapsto \bigg(1,\ x,\ x^2,\ x^3,\ (x-k_1)^3_+,\ (x-k_2)^3_+,\ldots, (x-k_p)^3_+\bigg)$ for some choices $\{k_1,\ldots,k_p\}$

Fourier series
: $x \mapsto \bigg(1,\ \cos(2\pi x),\ \sin(2\pi x),\ \cos(2\pi 2 x),\ \sin(2\pi 2 x), \ldots, \cos(2\pi p x),\ \sin(2\pi p x)\bigg)$



## How do you choose?

[Procedure 1:]{.secondary}

1. Pick your favorite basis. This is not as easy as it sounds. For instance, if $f$ is a step function, linear splines will do well with good knots, but polynomials will be terrible unless you have __lots__ of terms.

2. Perform OLS on different orders.

3. Use model selection criterion to choose the order.

[Procedure 2:]{.secondary}

1. Use a bunch of high-order bases, say Linear splines and Fourier series and whatever else you like.

2. Use Lasso or Ridge regression or elastic net. (combining bases can lead to multicollinearity, but we may not care)

3. Use model selection criteria to choose the tuning parameter.


## Try both procedures

1. Split `arcuate` into 75% training data and 25% testing data.

2. Estimate polynomials up to 20 as before and choose best order.

3. Do ridge, lasso and elastic net $\alpha=.5$ on 20th order polynomials, B splines with 20 knots, and Fourier series with $p=20$. Choose tuning parameter (using `lambda.1se`).

4. Repeat 1-3 10 times (different splits)


##


```{r simulation}
library(glmnet)
mapto01 <- function(x, pad = .005) (x - min(x) + pad) / (max(x) - min(x) + 2 * pad)
x <- mapto01(arcuate$position)
Xmat <- cbind(
  poly(x, 20), 
  splines::bs(x, df = 20), 
  cos(2 * pi * outer(x, 1:20)), sin(2 * pi * outer(x, 1:20))
)
y <- arcuate$fa
rmse <- function(z, s) sqrt(mean( (z - s)^2 ))
nzero <- function(x) with(x, nzero[match(lambda.1se, lambda)])
sim <- function(maxdeg = 20, train_frac = 0.75) {
  n <- nrow(arcuate)
  train <- as.logical(rbinom(n, 1, train_frac))
  test <- !train # not precisely 25%, but on average
  polycv <- map_dbl(seq(maxdeg), ~ cv_nice(lm(y ~ Xmat[,seq(.)], subset = train))) # figure out which order to use
  bpoly <- lm(y[train] ~ Xmat[train, seq(which.min(polycv))]) # now use it
  lasso <- cv.glmnet(Xmat[train, ], y[train])
  ridge <- cv.glmnet(Xmat[train, ], y[train], alpha = 0)
  elnet <- cv.glmnet(Xmat[train, ], y[train], alpha = .5)
  tibble(
    methods = c("poly", "lasso", "ridge", "elnet"),
    rmses = c(
      rmse(y[test], cbind(1, Xmat[test, 1:which.min(polycv)]) %*% coef(bpoly)),
      rmse(y[test], predict(lasso, Xmat[test,])),
      rmse(y[test], predict(ridge, Xmat[test,])),
      rmse(y[test], predict(elnet, Xmat[test,]))
    ),
    nvars = c(which.min(polycv), nzero(lasso), nzero(ridge), nzero(elnet))
  )
}
set.seed(12345)
sim_results <- map(seq(20), sim) |> list_rbind() # repeat it 20 times
```

## 

```{r sim-results}
#| code-fold: true
sim_results |>  
  pivot_longer(-methods) |> 
  ggplot(aes(methods, value, fill = methods)) + 
  geom_boxplot() +
  facet_wrap(~ name, scales = "free_y") + 
  ylab("") +
  theme(legend.position = "none") + 
  xlab("") +
  scale_fill_viridis_d(begin = .2, end = 1)
```


## Common elements

In all these cases, we transformed $x$ to a [higher-dimensional space]{.secondary}

Used $p+1$ dimensions with polynomials

Used $p+4$ dimensions with cubic splines

Used $2p+1$ dimensions with Fourier basis

## Featurization

Each case applied a [feature map]{.secondary} to $x$, call it $\Phi$

We used new "features" $\Phi(x) = \bigg(\phi_1(x),\ \phi_2(x),\ldots,\phi_k(x)\bigg)$

Neural networks (coming in module 4) use this idea

You've also probably seen it in earlier courses when you added interaction terms or other transformations.

. . .

Some methods (notably Support Vector Machines and Ridge regression) allow $k=\infty$

See [ISLR] 9.3.2 for baby overview or [ESL] 5.8 (note 😱)


# Next time...

Kernel regression and nearest neighbors
