---
lecture: "10 Basis expansions"
format: revealjs
metadata-files: 
  - _metadata.yml
---

{{< include _titleslide.qmd >}}

# Quick Review of Ridge and Lasso

## OLS: low bias, (potentially) high variance

$$
\begin{gathered}
\text{Model:} \quad y = x^\top \beta + \epsilon, \qquad \epsilon \sim N(0, \sigma^2)
\\
\text{OLS:} \quad \bls = \argmin_\beta \| \y - \X\beta\|_2^2\quad
\end{gathered}
$$

- Bias: $\E[\bls] - \beta = \E[\E[\bls \mid \X]] - \beta = \ldots = 0$
- variance: $\Var{\bls} =  \sigma^2(\X^\top \X)^{-1}$

[When is $(\X^\top \X)^{-1}$ large?]{.secondary}

. . . 

When we have *nearly colinear features*\
Nearly colinear features $\Rightarrow$ small singular values $\Rightarrow$ large matrix inverse\
[(Colinearity is more likely when $n$ is small)]{.secondary}


## Reducing variance (at the cost of additional bias)

0. *Manual variable selection*
1. *Ridge regression*: $\min_\beta \| \y - \X\beta\|_2^2 + \lambda \Vert \beta \Vert_2^2$
2. *Lasso*: $\min_\beta \| \y - \X\beta\|_2^2 + \lambda \Vert \beta \Vert_1$

. . .

*Ridge* shrinks all parameters towards 0.\
*Lasso* performs automatic variable selection.\

. . .

\
Increasing $\lambda$ *increases bias* and *decreases variance*.\
[(For ridge, larger lambda $\rightarrow$ smaller $\beta$.)]{.small}\
[(For lasso, larger lambda $\rightarrow$ sparser $\beta$.)]{.small}


---

```{r, fig.width=11,fig.align="center",dev="svg",fig.height=7, echo=FALSE}
library(glmnet)
data(prostate, package = "ElemStatLearn")
X <- prostate |> dplyr::select(-train, -lpsa) |>  as.matrix()
Y <- prostate$lpsa
ridge <- glmnet(X, Y, alpha = 0, lambda.min.ratio = 1e-10) # added to get a minimum
lasso <- glmnet(X, Y, alpha = 1, lambda.min.ratio = 1e-10) # added to get a minimum
ridge.cv <- cv.glmnet(X, Y, alpha = 0, lambda.min.ratio = 1e-10) # added to get a minimum
lasso.cv <- cv.glmnet(X, Y, alpha = 1, lambda.min.ratio = 1e-10) # added to get a minimum
par(mfrow = c(2, 2))
plot(ridge, main = "Ridge", xvar = "lambda")
plot(lasso, main = "Lasso", xvar = "lambda")
plot(ridge.cv, main = "Ridge", xvar = "lambda")
plot(lasso.cv, main = "Lasso", xvar = "lambda")
```


## Computing ridge and lasso predictors

- *OLS:* $\bls = (\X^\top \X)^{-1}\X^\top \y$
- *Ridge:* $\brl = (\X^\top \X + \lambda \mathbf{I})^{-1}\X^\top \y$
- *Lasso:* No closed form solution 😔
  - (Convex optimization problem solvable with iterative algorithms.)


## (Optional) Proof that ridge shrinks parameters

(This proof is not too hard if you use facts about SVDs and eigenvalues. I recommend working through it.)

Let $\mathbf{UDV^\top} = X$ be the SVD of $\mathbf X$.

$$
\begin{align}
\brl &= (\X^\top \X + \lambda \mathbf{I})^{-1} \X^\top \y
     \\
     &= (\X^\top \X + \lambda \mathbf{I})^{-1} {\color{blue} \X^\top \X (\X^\top \X)^{-1}} \X^\top \y
     \\
     &= (\X^\top \X + \lambda \mathbf{I})^{-1} \X^\top \X \underbrace{\left( (\X^\top \X)^{-1} \X^\top \y \right)}_{\bls}
     \\
     &= (\mathbf V \mathbf D^2 \mathbf V^\top + \lambda \mathbf{I})^{-1} \mathbf V \mathbf D^2 \mathbf V^\top \bls
     \\
     &= (\mathbf V (\mathbf D^2 + \lambda I) \mathbf V^\top)^{-1} \mathbf V \mathbf D^2 \mathbf V^\top \bls
     \\
     &= \mathbf V (\mathbf D^2)(\mathbf D^2 + \lambda I)^{-1} \mathbf V^\top \bls
\end{align}
$$

---

$(\mathbf D^2)(\mathbf D^2 + \lambda I)^{-1}$ is a diagonal matrix with entries $d_i^2/(d_i^2 + \lambda) < 1$.

So $\mathbf V (\mathbf D^2)(\mathbf D^2 + \lambda I)^{-1} \mathbf V^\top$ is a matrix that *shrinks* all coefficients of any vector multiplied against it.

$$
\Vert \mathbf V (\mathbf D^2)(\mathbf D^2 + \lambda I)^{-1} \mathbf V^\top \bls \Vert_2 < \Vert \bls \Vert_2.
$$
So $\Vert \brl \Vert_2 < \Vert \bls \Vert_2$


# Now onto new stuff
(But first, more clickers!)

## What about nonlinear things


$$\text{Our usual model:} \quad \Expect{Y \given X=x} = \sum_{j=1}^p x_j\beta_j$$

Now we relax this assumption of linearity:

$$\Expect{Y \given X=x} = f(x)$$

How do we estimate $f$?

. . . 

For this lecture, we use $x \in \R$ (1 dimensional)

Higher dimensions are possible, but complexity grows [exponentially]{.secondary}.

We'll see some special techniques for $x\in\R^p$ later this Module.


## Start simple

For any $f : \R \rightarrow [0,1]$

$$f(x) = f(x_0) + f'(x_0)(x-x_0) + \frac{1}{2}f''(x_0)(x-x_0)^2 + \frac{1}{3!}f'''(x_0)(x-x_0)^3 + R_3(x-x_0)$$

So we can linearly regress $y_i = f(x_i)$ on the polynomials.

The more terms we use, the smaller $R$.

```{r}
#| code-fold: true
set.seed(406406)
data(arcuate, package = "Stat406") 
arcuate <- arcuate |> slice_sample(n = 220)
arcuate %>% 
  ggplot(aes(position, fa)) + 
  geom_point(color = blue) +
  geom_smooth(color = orange, formula = y ~ poly(x, 3), method = "lm", se = FALSE)
```


## Same thing, different orders

```{r}
#| code-fold: true
arcuate %>% 
  ggplot(aes(position, fa)) + 
  geom_point(color = blue) + 
  geom_smooth(aes(color = "a"), formula = y ~ poly(x, 4), method = "lm", se = FALSE) +
  geom_smooth(aes(color = "b"), formula = y ~ poly(x, 7), method = "lm", se = FALSE) +
  geom_smooth(aes(color = "c"), formula = y ~ poly(x, 25), method = "lm", se = FALSE) +
  scale_color_manual(name = "Taylor order",
    values = c(green, red, orange), labels = c("4 terms", "7 terms", "25 terms"))
```


## Still a "linear smoother"

Really, this is still linear regression, just in a transformed space.

It's not linear in $x$, but it is linear in $(x,x^2,x^3)$ (for the 3rd-order case)

So, we're still doing OLS with

$$\X=\begin{bmatrix}1& x_1 & x_1^2 & x_1^3 \\ \vdots&&&\vdots\\1& x_n & x_n^2 & x_n^3\end{bmatrix}$$

So we can still use our nice formulas for LOO-CV, GCV, Cp, AIC, etc.


```{r}
max_deg <- 20
cv_nice <- function(mdl) mean( residuals(mdl)^2 / (1 - hatvalues(mdl))^2 ) 
cvscores <- map_dbl(seq_len(max_deg), ~ cv_nice(lm(fa ~ poly(position, .), data = arcuate)))
```

## 

```{r}
#| code-fold: true
library(cowplot)
g1 <- ggplot(tibble(cvscores, degrees = seq(max_deg)), aes(degrees, cvscores)) +
  geom_point(colour = blue) +
  geom_line(colour = blue) + 
  labs(ylab = 'LOO-CV', xlab = 'polynomial degree') +
  geom_vline(xintercept = which.min(cvscores), linetype = "dotted") 
g2 <- ggplot(arcuate, aes(position, fa)) + 
  geom_point(colour = blue) + 
  geom_smooth(
    colour = orange, 
    formula = y ~ poly(x, which.min(cvscores)), 
    method = "lm", 
    se = FALSE
  )
plot_grid(g1, g2, ncol = 2)
```


## Other bases

Polynomials
: $x \mapsto \left(1,\ x,\ x^2, \ldots, x^p\right)$ (technically, not quite this, they are orthogonalized)

Linear splines
: $x \mapsto \bigg(1,\ x,\ (x-k_1)_+,\ (x-k_2)_+,\ldots, (x-k_p)_+\bigg)$ for some $\{k_1,\ldots,k_p\}$

<!--
Cubic splines
: $x \mapsto \bigg(1,\ x,\ x^2,\ x^3,\ (x-k_1)^3_+,\ (x-k_2)^3_+,\ldots, (x-k_p)^3_+\bigg)$ for some choices $\{k_1,\ldots,k_p\}$
-->

Fourier series
: $x \mapsto \bigg(1,\ \cos(2\pi x),\ \sin(2\pi x),\ \cos(2\pi 2 x),\ \sin(2\pi 2 x), \ldots, \cos(2\pi p x),\ \sin(2\pi p x)\bigg)$

```{r fig.height=3, fig.width=9}
#| code-fold: true
library(cowplot)
library(ggplot2)

relu_shifted <- function(x, shift) {pmax(0, x - shift)}

# Create a sequence of x values
x_vals <- seq(-3, 3, length.out = 1000)

# Create a data frame with all the shifted functions
data <- data.frame(
  x = rep(x_vals, 5),
  polynomial = c(x_vals, x_vals^2, x_vals^3, x_vals^4, x_vals^5),
  linear.splines = c(relu_shifted(x_vals, 2), relu_shifted(x_vals, 1), relu_shifted(x_vals, 0), relu_shifted(x_vals, -1), relu_shifted(x_vals, -2)),
  fourier = c(cos(pi / 2 * x_vals), sin(pi / 2 * x_vals), cos(pi / 4 * x_vals), sin(pi / 4 * x_vals), cos(pi * x_vals)),
  function_label = rep(c("f1", "f2", "f3", "f4", "f5"), each = length(x_vals))
)

# Plot using ggplot2
g1 <- ggplot(data, aes(x = x, y = polynomial, color = function_label)) +
      geom_line(size = 1, show.legend=FALSE) +
      theme(axis.text.y=element_blank())
g2 <- ggplot(data, aes(x = x, y = linear.splines, color = function_label)) +
      geom_line(size = 1, show.legend=FALSE) +
      theme(axis.text.y=element_blank())
g3 <- ggplot(data, aes(x = x, y = fourier, color = function_label)) +
      geom_line(size = 1, show.legend=FALSE) +
      theme(axis.text.y=element_blank())

plot_grid(g1, g2, g3, ncol = 3)
```

## How do you choose?

[Procedure 1:]{.secondary}

1. Pick your favorite basis. (Think if the data might "prefer" one basis over another.)
    - [How "smooth" is the response you're trying to model?]{.small}
  
<!-- This is not as easy as it sounds. For instance, if $f$ is a step function, linear splines will do well with good knots, but polynomials will be terrible unless you have __lots__ of terms. -->

2. Perform OLS on different orders.

3. Use model selection criterion to choose the order.

---

```{r echo=FALSE, fig.height=2, fig.width=6}
plot_grid(g1, g2, g3, ncol = 3)
```

```{r fig.height=4, fig.width=12, echo=FALSE}
set.seed(12345)
coeff <- rnorm(5)
funcs <- data.frame(x = x_vals,
                    f1 = (matrix(data$polynomial, nrow = length(x_vals), ncol = 5)  %*% coeff),
                    f2 = (matrix(data$fourier, nrow = length(x_vals), ncol = 5)  %*% coeff),
                    f3 = (matrix(data$linear.splines, nrow = length(x_vals), ncol = 5)  %*% coeff))

g4 <- ggplot(funcs, aes(x=x, y=f1)) + geom_line(color = "blue") + theme()
g5 <- ggplot(funcs, aes(x=x, y=f2)) + geom_line(color = "blue") + theme()
g6 <- ggplot(funcs, aes(x=x, y=f3)) + geom_line(color = "blue") + theme()

plot_grid(g4, g5, g6, ncol = 3)
```

What bases do you think will work best for $f1$, $f2$ and $f3$?

. . .

[*Answer: $f1$ was made from polynomial bases, $f2$ from fourier, $f3$ from linear splines*]{.secondary}


---

## How do you choose?

[Procedure 2:]{.secondary}

1. Use a bunch of high-order bases, say Linear splines and Fourier series and whatever else you like.

2. Use Lasso or Ridge regression or elastic net. (combining bases can lead to multicollinearity, but we may not care)

3. Use model selection criteria to choose the tuning parameter.


## Try both procedures

1. Split `arcuate` into 75% training data and 25% testing data.

2. Estimate polynomials up to 20 as before and choose best order.

3. Do ridge, lasso and elastic net $\alpha=.5$ on 20th order polynomials, splines with 20 knots, and Fourier series with $p=20$. Choose tuning parameter (using `lambda.1se`).

4. Repeat 1-3 10 times (different splits)


##


```{r simulation}
library(glmnet)
mapto01 <- function(x, pad = .005) (x - min(x) + pad) / (max(x) - min(x) + 2 * pad)
x <- mapto01(arcuate$position)
Xmat <- cbind(
  poly(x, 20), 
  splines::bs(x, df = 20, degree = 1), 
  cos(2 * pi * outer(x, 1:20)), sin(2 * pi * outer(x, 1:20))
)
y <- arcuate$fa
rmse <- function(z, s) sqrt(mean( (z - s)^2 ))
nzero <- function(x) with(x, nzero[match(lambda.1se, lambda)])
sim <- function(maxdeg = 20, train_frac = 0.75) {
  n <- nrow(arcuate)
  train <- as.logical(rbinom(n, 1, train_frac))
  test <- !train # not precisely 25%, but on average
  polycv <- map_dbl(seq(maxdeg), ~ cv_nice(lm(y ~ Xmat[,seq(.)], subset = train))) # figure out which order to use
  bpoly <- lm(y[train] ~ Xmat[train, seq(which.min(polycv))]) # now use it
  lasso <- cv.glmnet(Xmat[train, ], y[train])
  ridge <- cv.glmnet(Xmat[train, ], y[train], alpha = 0)
  elnet <- cv.glmnet(Xmat[train, ], y[train], alpha = .5)
  tibble(
    methods = c("poly", "lasso", "ridge", "elnet"),
    rmses = c(
      rmse(y[test], cbind(1, Xmat[test, 1:which.min(polycv)]) %*% coef(bpoly)),
      rmse(y[test], predict(lasso, Xmat[test,])),
      rmse(y[test], predict(ridge, Xmat[test,])),
      rmse(y[test], predict(elnet, Xmat[test,]))
    ),
    nvars = c(which.min(polycv), nzero(lasso), nzero(ridge), nzero(elnet))
  )
}
set.seed(12345)
sim_results <- map(seq(20), sim) |> list_rbind() # repeat it 20 times
```

## 

```{r sim-results}
#| code-fold: true
sim_results |>  
  pivot_longer(-methods) |> 
  ggplot(aes(methods, value, fill = methods)) + 
  geom_boxplot() +
  facet_wrap(~ name, scales = "free_y") + 
  ylab("") +
  theme(legend.position = "none") + 
  xlab("") +
  scale_fill_viridis_d(begin = .2, end = 1)
```


## Common elements

In all these cases, we transformed $x$ to a [higher-dimensional space]{.secondary}

Used $p+1$ dimensions with polynomials

Used $p+4$ dimensions with cubic splines

Used $2p+1$ dimensions with Fourier basis

## Featurization

Each case applied a [feature map]{.secondary} to $x$, call it $\Phi$

We used new "features" $\Phi(x) = \bigg(\phi_1(x),\ \phi_2(x),\ldots,\phi_k(x)\bigg)$
w/ a linear model

$$f(x) = \Phi(x)^\top \beta$$ 

Neural networks (coming in module 4) build upon this idea

<!-- You've also probably seen it in earlier courses when you added interaction terms or other transformations. -->

. . .

\
Some methods (Support Vector Machines and other Kernel Machines) allow $k=\infty$\
[See [ISLR] 9.3.2 for baby overview or [ESL] 5.8 (note 😱)]{.small}


# Next time...

Kernel regression and nearest neighbors
