---
title: "00 Quiz 0 fun"
author: 
  - "STAT 406"
  - "Daniel J. McDonald"
date: 'Last modified - `r Sys.Date()`'
---


## Why this class?

* Most say requirements.
* Interest in ML/Stat learning
* Some cracks about preferring to code in other languages (not R)
* Expressions of love/affection for Stats/CS/ML
* Enjoyment of past similar classes

--

More idiosyncratic:

--

(Paraphrase) "Was going to drop since it's 8am + Clickers, but your website is cool"

--

"Learn strategies."

--
 
 "MACHINE LEARNING *DABS*"


---

## Syllabus Q

```{r setup, include=FALSE}
source("rmd_config.R")
theme_set(theme_bw(16))
quiz <- readRDS(here::here("..", "..", "WinterT1-2022", "quiz0", "quiz0-3.rds"))
```

```{r, echo=FALSE, fig.height=6, fig.align='center', fig.width=12}
quiz %>% 
  ggplot(aes(str_wrap(syllabus, 20))) + 
  geom_bar(fill = blue) + 
  scale_y_continuous(expand = expansion(c(0,0.05))) +
  xlab("") 
```

---

## Programming languages

```{r, echo=FALSE, fig.height=6, fig.width=12, fig.align='center'}
quiz %>% 
  ggplot(aes(str_wrap(R,20))) + geom_bar(fill=blue) + 
  scale_y_continuous(expand = expansion(c(0,0.05))) +
  xlab("")
library(MASS)
```

---

## Matrix inversion

```{r}
X = matrix(c(5,3,1,-1), nrow=2)
X
solve(X)
ginv(X)
X^(-1)
```

---

## Linear models

```{r}
y = X %*% c(2,-1) + rnorm(2)
coefficients(lm(y~X))
coef(lm(y~X))
solve(t(X) %*% X) %*% t(X) %*% y
# X \ y this it Matlab
```

---

## Pets and plans

```{r, echo=FALSE, fig.height=6, fig.width=12, fig.align='center'}
library(cowplot)
p1 = quiz %>% 
  ggplot(aes(str_wrap(pet,30))) + geom_bar(fill=blue) + 
  coord_flip() + 
  scale_y_continuous(expand = expansion(c(0,0.05))) +
  xlab("")
p2 = quiz %>% 
  ggplot(aes(str_wrap(post_grad,20))) + geom_bar(fill=orange) + 
  coord_flip() + 
  scale_y_continuous(expand = expansion(c(0,0.05))) +
  xlab("")
plot_grid(p1,p2)
```

---

## Grade predictions

```{r, echo=FALSE, fig.height=4, fig.width=12, fig.align='center'}
s406 = as.factor(rep(c("50-54%","55-59%","60-63%","64-67%","68-71%","72-75%","76-79%","80-84%","85-89%","90-100%"), times=c(0,0,3,7,4,18,28,24,20,11)))
preds = cut(quiz$final_grade[!is.na(quiz$final_grade)], 
            c(0,50,54,59,63,67,71,75,79,84,89,101),
            c("<50%","50-54%","55-59%","60-63%","64-67%","68-71%","72-75%","76-79%","80-84%","85-89%","90-100%"))
data.frame(name=rep(c("2021 distribution", "your predictions"), 
                    times=c(length(s406),length(preds))), 
           value=fct_c(s406,preds)) %>%
  mutate(value = fct_relevel(value, "<50%")) %>%
  ggplot(aes(value, fill=name)) + geom_bar(position = position_dodge()) +
  scale_y_continuous(expand = expansion(c(0,0.05))) +
  scale_fill_manual(values=c(blue,orange))
```

* 3 people say 100%
* 26 say 90%
* 36 say 85%
* 27 say 80%
* Lots of clumping


--

1 said 76.83, which is a really specific B+
