---
title: "00 R, Rmarkdown, code, and {tidyverse}: <br> A whirlwind tour"
author: 
  - "STAT 406"
  - "Daniel J. McDonald"
date: 'Last modified - `r Sys.Date()`'
---
class: middle, inverse, center


```{r setup, include=FALSE}
source("rmd_config.R")
library(magrittr)
```

```{r css-extras, file="css-extras.R", echo=FALSE}
```



# The basics

---
## Tour of Rstudio

Things to note

1. Console
2. Terminal
3. Scripts, .Rmd, Knit
4. Files, Projects
5. Getting help
6. Environment, Git

---
## Simple stuff

.pull-left[
### Vectors:
```{r}
x <- c(1, 3, 4)
x[1]
x[-1]
rev(x)
c(x, x)
```
]

.pull-right[
### Matrices:
```{r}
x <- matrix(1:25, nrow = 5, ncol = 5)
x[1,]
x[,-1]
x[c(1,3),  2:3]
```

]

---
## Simple stuff

.pull-left[
### Lists
```{r}
(l <- list(a = letters[1:2], b = 1:4, c = list(a = 1)))
l$a
l$c$a
l["b"] # compare to l[["b"]] == l$b
```
]

.pull-right[
### Data frames
```{r}
(dat <- data.frame(z = 1:5, b = 6:10, c = letters[1:5]))
class(dat)
dat$b
dat[1,]
```

.secondary[Data frames are sort-of lists and sort-of matrices]
]

---

## Tibbles

.secondary[These are `{tidyverse}` data frames]

```{r}
(dat2 <- tibble(z = 1:5, b = z + 5, c = letters[z]))
class(dat2)
```

We'll return to classes in a moment. A `tbl_df` is a "subclass" of `data.frame`.

Anything that `data.frame` can do, `tbl_df` can do (better).

For instance, the printing is more informative.

Also, you can construct one by referencing previously constructed columns.
---
class: middle, inverse, center

# Functions

---

## Understanding signatures

```{r echo=FALSE}
sig <- sig::sig
```

```{r}
sig(lm)
sig(`+`)
sig(ggplot2::layer)
sig(epipredict::layer)
sig(rnorm)
```

---

## These are all the same

```{r}
set.seed(12345)
rnorm(3)
set.seed(12345)
rnorm(n = 3, mean = 0)
set.seed(12345)
rnorm(3, 0, 1)
set.seed(12345)
rnorm(sd = 1, n = 3, mean = 0)
```

* Functions can have default values.
* You may, but don't have to, name the arguments
* If you name them, you can pass them out of order (but you shouldn't).

---

## Write lots of functions. I can't emphasize this enough.

.pull-left[
```{r functions}
f <- function(arg1, arg2, arg3 = 12, ...) {
  stuff <- arg1 * arg3
  stuff2 <- stuff + arg2
  plot(arg1, stuff2, ...)
  return(stuff2)
}
x <- rnorm(100)
```
]

.pull-right[
```{r plot-it}
y1 <- f(x, 3, 15, col = 4, pch = 19)
```
]

---

## Outputs vs. Side effects

.pull-left[
* Side effects are things a function does, outputs can be assigned to variables
* A good example is the `hist` function
* You have probably only seen the side effect which is to plot the histogram
```{r,fig.align='center', fig.height=5}
my_histogram <- hist(rnorm(1000))
```
]

.pull-right[
```{r}
my_histogram
class(my_histogram)
```

]

---

### When writing functions, program defensively, ensure behaviour

.pull-left[
```{r, error=TRUE}
incrementer <- function(x, inc_by = 1) return(x + 1)
incrementer(2)
incrementer(1:4)
incrementer("a")
```

```{r, error=TRUE}
incrementer <- function(x, inc_by = 1) {
  stopifnot(is.numeric(x))
  return(x + 1)
}
incrementer("a")
```

]

--

.pull-right[
```{r, error=TRUE}
incrementer <- function(x, inc_by = 1) {
  if (!is.numeric(x)) stop("`x` should be numeric")
  return(x + 1)
}
incrementer("a")
incrementer(2, -3) ## oops!
incrementer <- function(x, inc_by = 1) {
  if (!is.numeric(x)) stop("`x` should be numeric")
  return(x + inc_by)
}
incrementer(2, -3)
```
]

---

## How to keep track

.pull-left[
```{r}
library(testthat)
incrementer <- function(x, inc_by = 1) {
  if (!is.numeric(x)) stop("`x` should be numeric")
  if (!is.numeric(inc_by)) 
    stop("`inc_by` should be numeric")
  return(x + inc_by)
}
expect_error(incrementer("a"))
expect_equal(incrementer(1:3), 2:4)
expect_equal(incrementer(2, -3), -1)
expect_error(incrementer(1, "b"))
```

```{r, error=TRUE}
expect_identical(incrementer(1:3), 2:4)
```
]

--

.pull-right[
```{r}
is.integer(2:4)
is.integer(incrementer(1:3))
expect_identical(incrementer(1:3, 1L), 2:4)
```
]

--

.emphasis[
If you copy something, write a function.

Validate your arguments.

To ensure proper functionality, write tests to check if inputs result in predicted outputs.
]


---
class: middle, inverse, center

# Classes and methods

---

## Classes

.pull-left[
We saw some of these earlier:

```{r}
tib <- tibble(
  x1 = rnorm(100), 
  x2 = rnorm(100), 
  y = x1 + 2 * x2 + rnorm(100)
)
mdl <- lm(y ~ ., data = tib )
class(tib)
class(mdl)
```

The class allows for the use of "methods"

```{r}
print(mdl)
```
]

--

.pull-right[
`R` "knows what to do" when you `print()` an object of class `"lm"`.

`print()` is called a "generic" function. You can create "methods" that get dispatched.

For any generic, `R` looks for a method for the class.

If available, it calls that function.

```{r}

sloop::s3_dispatch(print(incrementer))
sloop::s3_dispatch(print(tib))
sloop::s3_dispatch(print(mdl))
```
]

---


## R-Geeky But Important

There are .secondary[lots] of generic functions in `R`

Common ones are `print()`, `summary()`, and `plot()`.

Also, lots of important statistical modelling concepts:
`residuals()` `coef()` 

(In `python`, these work the opposite way: `obj.residuals`. The dot after the object accesses methods defined for that type of object. But the dispatch behaviour is less robust.)  

* The convention is
that the specialized function is named `method.class`, e.g., `summary.lm`.

* If no specialized function is defined, R will try to use `method.default`.

For this reason, `R` programmers try to avoid `.` in names of functions or objects.

--

### Wherefore methods?


* The advantage is that you don't have to learn a totally
new syntax to grab residuals or plot things

* You just use `residuals(mdl)` whether `mdl` comes from `lm`
could have been done two centuries ago, or a Batrachian Emphasis Machine
which won't be invented for another five years.  

* The one draw-back is the help pages for the generic methods tend
to be pretty vague

* Compare `?summary` with `?summary.lm`.  


---
class: middle, inverse, center

# Environments 

---

## Different environments

* These are often tricky, but are very common.

* Most programming languages have this concept in one way or another.

* In `R` code run in the Console produces objects in the "Global environment"

* You can see what you create in the "Environment" tab.

* But there's lots of other stuff.

* Many packages are automatically loaded at startup, so you have access to the functions and data inside

For example `mean()`, `lm()`, `plot()`, `iris` (technically `iris` is lazy-loaded, meaning it's not in memory until you call it, but it is available)

--

<hr>

* Other packages require you to load them with `library(pkg)` before their functions are available.

* But, you can call those functions by prefixing the package name `ggplot2::ggplot()`.

* You can also access functions that the package developer didn't "export" for use with `:::` like `dplyr:::as_across_fn_call()`

???

That is all about accessing "objects in package environments"

---

## Other issues with environments

.pull-left[
As one might expect, functions create an environment inside the function.
```{r}
z <- 1
fun <- function(x) {
  z <- x
  print(z)
  invisible(z)
}
fun(14)
```

More non-trivial cases are `data-masking` environments.

```{r, error=TRUE}
tib <- tibble(
  x1 = rnorm(100), 
  x2 = rnorm(100), 
  y = x1 + 2 * x2
)
mdl <- lm(y ~ x2, data = tib)
x2
```
* `lm()` looks "inside" the `tib` to find `y` and `x2`
* The data variables are added to the `lm` environment
]

--

.pull-right[

.emphasis[
When Knit, `.Rmd` files run in their OWN environment.
]

They are run from top to bottom, with code chunks depending on previous

This makes them reproducible.

Jupyter notebooks don't do this.

Objects in your local environment are not available in the `.Rmd`

Objects in the `.Rmd` are not available locally.

.emphasis[
The most frequent error I see is:

* running chunks individually, 1-by-1, and it works
* Knitting, and it fails

The reason is almost always that the chunks refer to objects in the Environment that don't exist in the `.Rmd`
]
]

---
class: middle

.alert[
### This error also happens because:

* `library()` calls were made globally but not in the `.Rmd` so the packages aren't loaded

* paths to data or other objects are not relative to the `.Rmd` in your file system (they must be)


]


<br><br>

* Carefully keeping Labs / Assignments in their current location will help to avoid some of these.


---
class: middle, inverse, center

# Debugging

---

## How to fix code

* If you're using a function in a package, start with `?function` to see the help
    * Make sure you're calling the function correctly.
    * Try running the examples.
    * paste the error into Google (if you share the error on Slack, I often do this first)
    * Go to the package website if it exists, and browse around
    
* If your `.Rmd` won't Knit
    * Did you make the mistake on the last slide?
    * Did it Knit before? Then the bug is in whatever you added.
    * Did you never Knit it? Why not?
    * Call `rstudioapi::restartSession()`, then run the Chunks 1-by-1
    
* Adding `browser()`
    * Only useful with your own functions.
    * Open the script with the function, and add `browser()` to the code somewhere
    * Then call your function.
    * The execution will Stop where you added `browser()` and you'll have access to the local objects (and the global objects) to play around

---

## Reproducible examples

.emphasis[
Question I get on Slack that I hate:

"I ran the code like you had on Slide 39, but it didn't work."
]

If you want to ask me why the code doesn't work, you need to show me what's wrong.

.emphasis[
**Don't just paste a screenshot!** 

Unless you get lucky, I won't be able to figure it out from that. And we'll both get frustrated.
]

What you need is a Reproducible Example or `reprex`.

This is a small chunk of code that runs in it's own environment and produces the error.

The best way to do this is with the `{reprex}` package.

---
## Reproducible examples, How it works

1. Open a new `.R` script.

1. Paste your buggy code in the file (no need to save)

1. Edit your code to make sure it's "enough to produce the error" and nothing more. (By rerunning the code a few times.)

1. Copy your code.

1. Call `reprex::reprex(venue = "r")` from the console. This will run your code in a new environment and show the result in the Viewer tab. Does it create the error you expect?

1. If it creates other errors, that may be the problem. You may fix the bug on your own!

1. If it doesn't have errors, then your global environment is Farblunget.

1. The Output is now on your clipboard. Go to Slack and paste it in a message. Then press `Cmd+Shift+Enter` (on Mac) or `Ctrl+Shift+Enter` (Windows/Linux). Under Type, select `R`.

1. Send the message, perhaps with more description and an SOS emoji.

.emphasis[
Note: Because Reprex runs in it's own environment, it doesn't have access to any of the libraries you loaded or the stuff in your global environment. You'll have to load these things in the script.
]

---
class: middle, inverse, center

# Understanding `{tidyverse}`

---

## `{tidyverse}` is huge

Core tidyverse is nearly 30 different R packages, but we're going to just talk about a few of them.

Falls roughly into a few categories:

1. **Convenience functions:** `{magrittr}` and many many others.
1. **Data processing:** `{dplyr}` and many others.
1. **Graphing:** `{ggplot2}` and some others like `{scales}`.
1. **Utilities**


--
<hr>

We're going to talk quickly about some of it, but ignore much of 2.

There's a lot that's great about these packages, especially ease of data processing.

But it doesn't always jive with base R (it's almost a separate proglang at this point).

---

## Piping

This was introduced by `{magrittr}` as `%>%`, but is now in base R (>=4.1.0) as `|>`.

Note: there are other pipes in `{magrittr}` (e.g. `%$%` and `%T%`) but I've never used them.

I've used the old version for so long, that it's hard for me to adopt the new one.

The point of the pipe is to **logically sequence nested operations**

--
<hr>

.pull-left[
```{r}
mse1 <- print(
  sum(
    residuals(
      lm(y~., data = mutate(
        tib, 
        x3 = x1^2,
        x4 = log(x2 + abs(min(x2)) + 1)
      )
      )
    )^2
  )
)
```
]

.pull-right[
```{r}
mse2 <- tib %>%
  mutate(
    x3 = x1^2, 
    x4 = log(x2 + abs(min(x2)) + 1)
  ) %>%
  lm(y ~ ., data = .) %>% ## note the use of `.`
  residuals() %>%
  raise_to_power(2) %>% ## same as `^`(2)
  sum() %>%
  print()
```
]

---

It may seem like we should push this all the way

```{r}
tib %>%
  mutate(
    x3 = x1^2, 
    x4 = log(x2 + abs(min(x2)) + 1)
  ) %>%
  lm(y ~ ., data = .) %>% ## note the use of `.`
  residuals() %>%
  raise_to_power(2) %>% ## same as `^`(2)
  sum() ->
  mse3
```

This works, but it's really annoying.

---

## Data processing in `{dplyr}`

.pull-left[
This package has all sorts of things. And it interacts with `{tibble}` generally.

The basic idea is "tibble in, tibble out".

Satisfies **data masking** which means you can refer to columns by name or use helpers like `ends_with("_rate")`

Majorly useful operations:

1. `select()` (chooses columns to keep)
1. `mutate()` (showed this already)
1. `group_by()`
1. `pivot_longer()` and `pivot_wider()`
1. `left_join()` and `full_join()`
1. `summarise()`

.emphasis[
Note that `filter()` and `select()` are functions in Base R. Sometimes you get `r fa("bug")` because it called the wrong version. To be sure, prefix it like `dplyr::select()`.
]
]

.pull-right[
A useful data frame
```{r}
library(epidatr)
covid <- covidcast(
  data_source = "jhu-csse",
  signals = "confirmed_7dav_incidence_prop,deaths_7dav_incidence_prop",
  time_type = "day",
  geo_type = "state",
  time_values = epirange(20220801, 20220821),
  geo_values = "ca,wa"
) %>%
  fetch_tbl() %>%
  select(geo_value, time_value, signal, value)
covid
```

]

---

.pull-left[

Rename the `signal` to something short.

```{r}
covid <- covid %>% 
  mutate(
    signal = case_when(
      str_starts(signal, "confirmed") ~ "case_rate",
      TRUE ~ "death_rate"
    ))
```


Sort by `time_value` then `geo_value`

```{r}
covid <- covid %>%
  arrange(time_value, geo_value)
```

Calculate grouped medians

```{r}
covid %>% 
  group_by(geo_value, signal) %>%
  summarise(med = median(value), .groups = "drop")
```

]

.pull-right[
Split the data into two tibbles by signal

```{r}
cases <- covid %>% 
  filter(signal == "case_rate") %>%
  rename(case_rate = value) %>% select(-signal)
deaths <- covid %>% 
  filter(signal == "death_rate") %>%
  rename(death_rate = value) %>% select(-signal)
```

Join them together
```{r}
joined <- full_join(
  cases, deaths, 
  by = c("geo_value", "time_value"))
```

Do the same thing by pivoting
```{r}
covid %>% 
  pivot_wider(names_from = signal, values_from = value)
```
]

---

## Plotting with `{ggplot2}`

* Everything you can do with `ggplot()`, you can do with `plot()`. But the 
defaults are _much_ prettier.

* It's also much easier to adjust by aesthetics/panels by factors.

* It also uses "data masking": data goes into `ggplot(data = mydata)`, then the columns are available to the rest.

* It (sort of) pipes, but by adding **layers** with `+`

* It .secondary[.hand[strongly prefers]] "long" data frames over "wide" data frames.

--

<hr>

I'll give a very fast overview of some confusing bits.

<br><br>

.center[
.secondary[.hand[I suggest exploring]]

`r fa("link")` .hand[[This slide deck](https://djnavarro.net/slides-starting-ggplot/)] `r fa("slideshare")`

.secondary[.hand[for more help]]
]

---
class: middle

.pull-left[

```{r adding-geoms, fig.show='hide'}
ggplot(
  data = covid %>% 
    filter(signal == "case_rate")
) +
  geom_point(
    mapping = aes(
      x = time_value,
      y = value
    )
  ) + 
  geom_smooth( 
    mapping = aes( 
      x = time_value, 
      y = value 
    ) 
  ) 
```
]

.pull-right[

```{r ref.label="adding-geoms", echo=FALSE, warning=FALSE, message=FALSE, fig.height=4}
```
]

???

* The complete code
* Data is specified in the ggplot, passed along
* (we show only case_rate)


* The Grey SE shading is pretty ugly
* And there are two states mashed together
* That trend is awfully wiggly


---
class: middle

.pull-left[

```{r adding-geoms2, fig.show='hide'}
ggplot(
  data = covid %>% 
    filter(signal == "case_rate")
) +
  geom_point(
    mapping = aes(
      x = time_value,
      y = value,
      colour = geo_value
    )
  ) + 
  geom_smooth( 
    mapping = aes( 
      x = time_value, 
      y = value,
      colour = geo_value
    ),
    se = FALSE,
    method = "lm"
  ) 
```
]

.pull-right[

```{r ref.label="adding-geoms2", echo=FALSE, warning=FALSE, message=FALSE, fig.height=4}
```
]

???

* Separate out the states by colour
* straight lines instead
* no more grey shading
* Why do I keep writing all that mapping = stuff?


---
class: middle

.pull-left[

```{r adding-geoms3, fig.show='hide'}
ggplot(
  data = covid %>% 
    filter(signal == "case_rate"),
  mapping = aes(
    x = time_value,
    y = value,
    colour = geo_value
  )
) +
  geom_point() + 
  geom_smooth(se = FALSE, method = "lm") 
```
]

.pull-right[

```{r ref.label="adding-geoms3", echo=FALSE, warning=FALSE, message=FALSE, fig.height=4}
```
]

???

mapping in the `ggplot()` call is shared across the rest

---
class: middle

.pull-left[

```{r adding-geoms4, fig.show='hide'}
ggplot(covid %>% filter(signal == "case_rate"),
       aes(time_value, value, colour = geo_value)) +
  geom_point() + 
  geom_smooth(se = FALSE, method = "lm") 
```
]

.pull-right[

```{r ref.label="adding-geoms4", echo=FALSE, warning=FALSE, message=FALSE, fig.height=4}
```
]

???

Don't need to name the arguments.

This is typically what ggplot code looks like.

Let's go a bit further to spruce this up.

---
class: middle

.pull-left[

```{r adding-geoms5, fig.show='hide'}
ggplot(covid, 
       aes(time_value, value, colour = geo_value)) +
  geom_point() + 
  geom_smooth(se = FALSE, method = "lm") +
  facet_grid(signal ~ geo_value) +
  scale_colour_manual(
    name = NULL, 
    values = c(blue, orange)) +
  theme(legend.position = "bottom")
```
]

.pull-right[

```{r ref.label="adding-geoms5", echo=FALSE, warning=FALSE, message=FALSE, fig.height=6}
```
]

???

* use facet_grid to split out states / show both signals (formula)
* change the colour scaling, remove the annoying title
* put the legend on the bottom
* But the y-axis scale is shared, measurements are on different scales

---
class: middle

.pull-left[

```{r adding-geoms6, fig.show='hide'}
ggplot(covid, 
       aes(time_value, value, colour = geo_value)) +
  geom_point() + 
  geom_smooth(se = FALSE, method = "lm") +
  facet_grid(signal ~ geo_value, scales = "free_y") +
  scale_colour_manual(
    name = NULL, 
    values = c(blue, orange)) +
  theme(legend.position = "bottom")
```
]

.pull-right[

```{r ref.label="adding-geoms6", echo=FALSE, warning=FALSE, message=FALSE, fig.height=6}
```
]