---
title: "Gradient Ascent: Logistic Regression"
author: DJM
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  learnr::tutorial:
    progressive: true
    theme: journal
    highlight: kate
    ace_theme: solarized_dark
    includes:
      in_header: !expr system.file("tutorials/tutorials-css.html",package="UBCstat406labs")
runtime: shiny_prerendered
---

<!--
Derived from IU Stat 432 ic9.Rmd
css: "/css/tutorials.css"
-->

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
s301gradedist <- UBCstat406labs::s301gradedist

tutorial_options(exercise.timelimit = 5, exercise.checker = gradethis::grade_learnr)

knitr::opts_chunk$set(echo = FALSE)
library(knitr)
library(tidyverse)
library(cowplot)
```

## Logistic Regression

###

Suppose $Y=1$ with probability $p(x)$ and $Y=0$ with probability $1-p(x)$. I want to model $P(Y=1| X=x)$. I'll assume that $p(x)/(1-p(x)) = ax$ for some scalar $a$. 

We're going to estimate $a$ given data.

### Generate Data

```{r generate-data, exercise=TRUE}
set.seed(20200405)
n = 100
a = 2
x = runif(n)*10-5
logit <- function(x) exp(x)/(1+exp(x))
p = logit(a*x)
df = tibble(x=x, y = rbinom(n, 1, p))
```

Now that the data is generated, let's go ahead and take a look at it.

```{r plotdata-setup}
set.seed(20200405)
n = 100
a = 2
x = runif(n)*10-5
logit <- function(x) exp(x)/(1+exp(x))
p = logit(a*x)
y = rbinom(n, 1, p)
df = tibble(x=x, y=y)
```

```{r plotdata, exercise=TRUE}
ggplot(df, aes(x,y)) + geom_point(color="red") +
  stat_function(fun=function(x) logit(a*x)) + theme_cowplot(14)
```

### The likelihood is given by
\[
L(y | a, x) = \prod_{i=1}^n p(x_i)^{y_i}(1-p(x_i))^{1-y_i}
\]

(Simple) gradient ascent (to maximize $L(a)$) is:

1. Input $a_0, \gamma>0, j_\max, \epsilon>0, \nabla L(a)$.
2. For $j=1,2,\ldots,j_\max$,
\[
a_j = a_{j-1} + \gamma \nabla L(a_{j-1})
\]
3. Stop if $\epsilon > |a_j - a_{j-1}|$.


### Write a function to find $a_{mle}$

Note that on the log scale, $\nabla L(a) = \sum (y_i - p_i) x_i$ where $p_i$ is as above. Write a function that will keep and store the entire chain of $a_{mle}$'s.

```{r amlefun-setup}
set.seed(20200405)
n = 100
a = 2
x = runif(n)*10-5
logit <- function(x) exp(x)/(1+exp(x))
p = logit(a*x)
y = rbinom(n, 1, p)
df = tibble(x=x, y=y)
```


```{r amlefun, exercise=TRUE}
amle <- function(x, y, a0, gam=0.5, jmax=100, eps=1e-6){
  err = 1e8
  j = 1
  a = double(jmax)
  a[1] = a0
  while(j < jmax && err>eps){
    j    = ____________
    p    = ____________
    ell  = ____________
    a[j] = ____________
    err = abs(a[j] - a[j-1])
  }
  a[1:j]
}

```

###

If done correctly, your code should look something like:

```{r amlecorrect,echo=TRUE}
amle <- function(x, y, a0, gam=0.5, jmax=100, eps=1e-6){
  err = 1e8
  j = 1
  a = double(jmax)
  a[1] = a0
  while(j < jmax && err>eps){
    j = j+1
    p = logit(a[j-1]*x)
    ell = sum((y-p)*x)
    a[j] = a[j-1] + gam * ell
    err = abs(a[j] - a[j-1])
  }
  a[1:j]
}
```

### Run your function and report the results

```{r ourmle1-setup}
set.seed(20200405)
n = 100
a = 2
x = runif(n)*10-5
logit <- function(x) exp(x)/(1+exp(x))
p = logit(a*x)
y = rbinom(n, 1, p)
df = tibble(x=x, y=y)
amle <- function(x, y, a0, gam=0.5, jmax=100, eps=1e-6){
  err = 1e8
  j = 1
  a = double(jmax)
  a[1] = a0
  while(j < jmax && err>eps){
    j = j+1
    p = logit(a[j-1]*x)
    ell = sum((y-p)*x)
    a[j] = a[j-1] + gam * ell
    err = abs(a[j] - a[j-1])
  }
  a[1:j]
}
```

Let's start by running the function with the default intputs and an initialization of $a_0$=5.

```{r ourmle1, exercise=TRUE}
amle(x, y, 5)
```


```{r ourmle2-setup}
set.seed(20200405)
n = 100
a = 2
x = runif(n)*10-5
logit <- function(x) exp(x)/(1+exp(x))
p = logit(a*x)
y = rbinom(n, 1, p)
df = tibble(x=x, y=y)
amle <- function(x, y, a0, gam=0.5, jmax=100, eps=1e-6){
  err = 1e8
  j = 1
  a = double(jmax)
  a[1] = a0
  while(j < jmax && err>eps){
    j = j+1
    p = logit(a[j-1]*x)
    ell = sum((y-p)*x)
    a[j] = a[j-1] + gam * ell
    err = abs(a[j] - a[j-1])
  }
  a[1:j]
}
```

Next, let's try changing the initialization to $a_0$=.1.

```{r ourmle2, exercise=TRUE}
amle(x, y, .1)
```


```{r ourmle3-setup}
set.seed(20200405)
n = 100
a = 2
x = runif(n)*10-5
logit <- function(x) exp(x)/(1+exp(x))
p = logit(a*x)
y = rbinom(n, 1, p)
df = tibble(x=x, y=y)
amle <- function(x, y, a0, gam=0.5, jmax=100, eps=1e-6){
  err = 1e8
  j = 1
  a = double(jmax)
  a[1] = a0
  while(j < jmax && err>eps){
    j = j+1
    p = logit(a[j-1]*x)
    ell = sum((y-p)*x)
    a[j] = a[j-1] + gam * ell
    err = abs(a[j] - a[j-1])
  }
  a[1:j]
}
```

Third, let's try changing the initialization back to $a_0$=5 but set $gam=.1$.

```{r ourmle3, exercise=TRUE}
amle(x, y, 5, .1)
```


```{r ourmle4-setup}
set.seed(20200405)
n = 100
a = 2
x = runif(n)*10-5
logit <- function(x) exp(x)/(1+exp(x))
p = logit(a*x)
y = rbinom(n, 1, p)
df = tibble(x=x, y=y)
amle <- function(x, y, a0, gam=0.5, jmax=100, eps=1e-6){
  err = 1e8
  j = 1
  a = double(jmax)
  a[1] = a0
  while(j < jmax && err>eps){
    j = j+1
    p = logit(a[j-1]*x)
    ell = sum((y-p)*x)
    a[j] = a[j-1] + gam * ell
    err = abs(a[j] - a[j-1])
  }
  a[1:j]
}
```

Finally, let's leave the initialization at $a_0$=5 but set $gam=1$.

```{r ourmle4, exercise=TRUE}
amle(x, y, 5, 1)
```

```{r ourmle5-setup}
set.seed(20200405)
n = 100
a = 2
x = runif(n)*10-5
logit <- function(x) exp(x)/(1+exp(x))
p = logit(a*x)
y = rbinom(n, 1, p)
df = tibble(x=x, y=y)
amle <- function(x, y, a0, gam=0.5, jmax=100, eps=1e-6){
  err = 1e8
  j = 1
  a = double(jmax)
  a[1] = a0
  while(j < jmax && err>eps){
    j = j+1
    p = logit(a[j-1]*x)
    ell = sum((y-p)*x)
    a[j] = a[j-1] + gam * ell
    err = abs(a[j] - a[j-1])
  }
  a[1:j]
}
```

We can compute the actual value of $a_{mle}$ by using the `glm()` function.  What value does this return?

```{r ourmle5, exercise=TRUE}
glm(y~x-1, family=binomial)$coef #just to check
```

###

```{r gen1-quiz}
quiz(caption="Quiz",
  question("Based on the output from `glm()`, which of the inputs returned the correct estimate of $a_{mle}$?",
    answer("`amle(x, y, 5)`"),
    answer("`amle(x, y, .1)`"),
    answer("`amle(x, y, 5, .1)`",correct=TRUE),
    answer("`amle(x, y, 5, 1)`"),
    allow_retry = TRUE
  )
)
```
