---
title: "Bootstrapping"
author: DJM
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  learnr::tutorial:
    progressive: true
    theme: journal
    highlight: kate
    ace_theme: solarized_dark
    includes:
      in_header: !expr system.file("tutorials/tutorials-css.html",package="UBCstat406labs")
runtime: shiny_prerendered
---

<!--
Derived from IU Stat 432 ic7.Rmd
css: "/css/tutorials.css"
-->

```{r setup, include=FALSE}
library(learnr)
library(gradethis)

tutorial_options(exercise.timelimit = 5, exercise.checker = gradethis::grade_learnr)

knitr::opts_chunk$set(echo = FALSE)
```


## Generate the Data

```{r generate-data-and-fit, exercise=TRUE, exercise.lines = 12}
## Note: this is the same data as in Lecture 8.
trueFunction <- function(x) sin(x) + 1/sqrt(x) + 3
set.seed(1234)
n = 100
x = 1:n/n*2*pi
wiggly = data.frame(x = x,
                 y = trueFunction(x) + rnorm(n, 0, .75))
library(ggplot2)
ggplot(wiggly, aes(x, y)) + geom_point() + xlim(0,2*pi) + ylim(0,max(wiggly$y)) +
  stat_function(fun=trueFunction, color='red')
```


## Function 1: Linear Model

Write a function that estimates the linear model of $y$ on $x$ for any data set. Call your function `wiggly.estimator.lm`. It should return the slope and intercept. 

```{r estimator-setup}
## Note: this is the same data as in Lecture 8.
trueFunction <- function(x) sin(x) + 1/sqrt(x) + 3
set.seed(1234)
n = 100
x = 1:n/n*2*pi
wiggly = data.frame(x = x,
                 y = trueFunction(x) + rnorm(n, 0, .75))
```

```{r estimator, exercise = TRUE, exercise.lines = 6}
wiggly.estimator.lm <- function(newdata=wiggly){
    lm.coef = _________________________
    lm.coef
} 
wiggly.estimator.lm()
```

```{r estimator-check}
trueFunction <- function(x) sin(x) + 1/sqrt(x) + 3
set.seed(1234)
n = 100
x = 1:n/n*2*pi
wiggly = data.frame(x = x,
                 y = trueFunction(x) + rnorm(n, 0, .75))
wiggly.estimator.lm <- function(newdata){
    lm.coef = coef(lm(y~x, data=newdata))
    lm.coef
} 
sol <- wiggly.estimator.lm(wiggly)


grade_result(
  pass_if(~ identical(.result, sol), "Correct!"),
  fail_if(~ TRUE, "Incorrect.")
)
```

## Function 2: Residual Resampling

###

Write a function that computes the residuals, resamples from those residuals, and then outputs a data.frame with the original $x$'s but new $y$'s.  Call it `wiggly.resids.resamp`.

```{r residsampler-setup}
trueFunction <- function(x) sin(x) + 1/sqrt(x) + 3
set.seed(1234)
n = 100
x = 1:n/n*2*pi
wiggly = data.frame(x = x,
                 y = trueFunction(x) + rnorm(n, 0, .75))

```

```{r residsampler, exercise=TRUE, exercise.lines=12}
wiggly.resids.resamp <- function(newdata=wiggly){
  newdata.ml = lm(y~x, data=newdata)
  resids    = __________________ # compute the residuals from the linear model
  newResids = __________________ # resample the residuals from the original model
  new.wiggles = data.frame( # create a new dataframe 
    x = ______________, # with the original x's
    y = fitted(newdata.ml)+newResids)  # but new y's
  return(new.wiggles)
}
set.seed(1234)
wiggly.resids.resamp()
```

```{r residsampler-check}
trueFunction <- function(x) sin(x) + 1/sqrt(x) + 3
set.seed(1234)
n = 100
x = 1:n/n*2*pi
wiggly = data.frame(x = x,
                 y = trueFunction(x) + rnorm(n, 0, .75))

wiggly.resids.resamp <- function(newdata=wiggly){
  newdata.ml = lm(y~x, data=newdata)
  resids = residuals(newdata.ml)
  newResids = sample(resids, replace=TRUE) # resample the residuals from the original model
  new.wiggles = data.frame( # create a new dataframe 
    x = newdata$x, # with the original x's but new y's
    y = fitted(newdata.ml)+newResids) 
  return(new.wiggles)
}
set.seed(1234)
wiggles = wiggly.resids.resamp(wiggly)

grade_result(
  pass_if(~ identical(.result, wiggles), "Correct!"),
  fail_if(~ TRUE, "Incorrect.")
)
```

###
Check your code with the following.  They should be similar.

```{r resid-sampler,echo=TRUE}
wiggly.resids.resamp <- function(newdata=wiggly){
  newdata.ml = lm(y~x, data=newdata)
  resids = residuals(newdata.ml)
  newResids = sample(resids, replace=TRUE) # resample the residuals from the original model
  new.wiggles = data.frame( # create a new dataframe 
    x = wiggly$x, # with the original x's but new y's
    y = fitted(newdata.ml)+newResids) 
  return(new.wiggles)
}
```

## Function 3: Resample Data.Frame

Write a function called `new.wiggly.resamp` to resample the data frame.  The following functions from your book can be referenced: `resample()`, `resample.data.frame()`, `rboot()`, `bootstrap()`, `equitails()`, `bootstrap.ci()`.

Hint: To write this function, it can be as easy as calling one of the functions above.

```{r dataframesampler-setup, echo=FALSE}
resample <- function(x) {
    sample(x, replace = TRUE)
}
resample.data.frame <- function(data) {
    sample.rows <- resample(1:nrow(data))
    return(data[sample.rows, ])
}
rboot <- function(statistic, simulator, B) {
    tboots <- replicate(B, statistic(simulator()))
    if (is.null(dim(tboots))) {
        tboots <- array(tboots, dim = c(1, B))
    }
    return(tboots)
}
bootstrap <- function(tboots, summarizer, ...) {
    summaries <- apply(tboots, 1, summarizer, ...)
    return(t(summaries))
}
equitails <- function(x, alpha) {
    lower <- quantile(x, alpha/2)
    upper <- quantile(x, 1 - alpha/2)
    return(c(lower, upper))
}
bootstrap.ci <- function(statistic = NULL, simulator = NULL, tboots = NULL,
    B = if (!is.null(tboots)) {
        ncol(tboots)
    }, t.hat, level) {
    if (is.null(tboots)) {
        stopifnot(!is.null(statistic))
        stopifnot(!is.null(simulator))
        stopifnot(!is.null(B))
        tboots <- rboot(statistic, simulator, B)
    }
    alpha <- 1 - level
    intervals <- bootstrap(tboots, summarizer = equitails, alpha = alpha)
    upper <- t.hat + (t.hat - intervals[, 1])
    lower <- t.hat + (t.hat - intervals[, 2])
    CIs <- cbind(lower = lower, upper = upper)
    return(CIs)
}
trueFunction <- function(x) sin(x) + 1/sqrt(x) + 3
set.seed(1234)
n = 100
x = 1:n/n*2*pi
wiggly = data.frame(x = x,
                 y = trueFunction(x) + rnorm(n, 0, .75))
```

```{r dataframesampler, exercise=TRUE, exercise.lines=4}
new.wiggly.resamp <- function(newdata=wiggly){
  __________________________ #resample the data frame
} 
new.wiggly.resamp()
```

```{r dataframesampler-check}
resample <- function(x) {
    sample(x, replace = TRUE)
}
resample.data.frame <- function(data) {
    sample.rows <- resample(1:nrow(data))
    return(data[sample.rows, ])
}
trueFunction <- function(x) sin(x) + 1/sqrt(x) + 3
set.seed(1234)
n = 100
x = 1:n/n*2*pi
wiggly = data.frame(x = x,
                 y = trueFunction(x) + rnorm(n, 0, .75))
new.wiggly.resamp <- function(newdata){
  resample.data.frame(newdata) #resample the data frame
} 
wiggles = new.wiggly.resamp(wiggly)

grade_result(
  pass_if(~ identical(.result, wiggles), "Correct!"),
  fail_if(~ TRUE, "Incorrect.")
)
```

## The Bootstrap

###

Use both samplers to get model-based and nonparametric confidence intervals for your linear model.

```{r bootstrap-setup}
resample <- function(x) {
    sample(x, replace = TRUE)
}
resample.data.frame <- function(data) {
    sample.rows <- resample(1:nrow(data))
    return(data[sample.rows, ])
}
rboot <- function(statistic, simulator, B) {
    tboots <- replicate(B, statistic(simulator()))
    if (is.null(dim(tboots))) {
        tboots <- array(tboots, dim = c(1, B))
    }
    return(tboots)
}
bootstrap <- function(tboots, summarizer, ...) {
    summaries <- apply(tboots, 1, summarizer, ...)
    return(t(summaries))
}
equitails <- function(x, alpha) {
    lower <- quantile(x, alpha/2)
    upper <- quantile(x, 1 - alpha/2)
    return(c(lower, upper))
}
bootstrap.ci <- function(statistic = NULL, simulator = NULL, tboots = NULL,
    B = if (!is.null(tboots)) {
        ncol(tboots)
    }, t.hat, level) {
    if (is.null(tboots)) {
        stopifnot(!is.null(statistic))
        stopifnot(!is.null(simulator))
        stopifnot(!is.null(B))
        tboots <- rboot(statistic, simulator, B)
    }
    alpha <- 1 - level
    intervals <- bootstrap(tboots, summarizer = equitails, alpha = alpha)
    upper <- t.hat + (t.hat - intervals[, 1])
    lower <- t.hat + (t.hat - intervals[, 2])
    CIs <- cbind(lower = lower, upper = upper)
    return(CIs)
}

trueFunction <- function(x) sin(x) + 1/sqrt(x) + 3
set.seed(1234)
n = 100
x = 1:n/n*2*pi
wiggly = data.frame(x = x,
                 y = trueFunction(x) + rnorm(n, 0, .75))
wiggly.estimator.lm <- function(newdata=wiggly){
    lm.coef = coef(lm(y~x, data=newdata))
    lm.coef
} 
wiggly.resids.resamp <- function(newdata=wiggly){
  newdata.ml = lm(y~x, data=newdata)
  resids = residuals(newdata.ml)
  newResids = sample(resids, replace=TRUE) # resample the residuals from the original model
  new.wiggles = data.frame( # create a new dataframe 
    x = wiggly$x, # with the original x's but new y's
    y = fitted(newdata.ml)+newResids) 
  return(new.wiggles)
}
new.wiggly.resamp <- function(newdata=wiggly){
  resample.data.frame(newdata) #resample the data frame
} 

```

```{r bootstrap, exercise=TRUE, exercise.lines=13}
mbb.wiggly = bootstrap.ci(
  statistic = wiggly.estimator.lm,
  simulator = wiggly.resids.resamp,
  B = 1000, t.hat = wiggly.estimator.lm(wiggly),
  level = 0.95)
npb.wiggly = bootstrap.ci(
  statistic = wiggly.estimator.lm,
  simulator = new.wiggly.resamp,
  B = 1000, t.hat = wiggly.estimator.lm(wiggly),
  level = 0.95)
mbb.wiggly
npb.wiggly
```

### Quiz

Which bootstrap is more appropriate to use in this case? Why?

###

Answer: The model-based bootstrap is wrong. Data did not actually come from this linear model. So resampling rows is a more accurate measure of uncertainty.







