---
title: "22 Neural nets - other considerations"
author:
- STAT 406
- Daniel J. McDonald
date: "Last modified - `r Sys.Date()`"
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
source("rmd_config.R")
```

## Estimation procedures (training)

$$\newcommand{\Expect}[1]{\mathbb{E}\left[ #1 \right]}
\newcommand{\Var}[1]{\mathbb{V}\left[ #1 \right]}
\newcommand{\Cov}[2]{\mathrm{Cov}\left[#1,\ #2\right]}
\newcommand{\given}{\ \vert\ }
\newcommand{\argmin}{\arg\min}
\newcommand{\argmax}{\arg\max}
\newcommand{\R}{\mathbb{R}}
\newcommand{\P}{\mathbb{P}}
\renewcommand{\hat}{\widehat}
\newcommand{\tr}[1]{\mbox{tr}(#1)}
\newcommand{\X}{\mathbf{X}}
\newcommand{\bfy}{\mathbf{y}}
\newcommand{\bfx}{\mathbf{x}}$$

Back-propagation

__Advantages:__ 

-   It's updates only depend on local
    information in the sense that if objects in the hierarchical model
    are unrelated to each other, the updates aren't affected

    (This helps in many ways, most notably in parallel architectures)

-   It doesn't require second-derivative information

-   As the updates are only in terms of $\hat{R}_i$, the algorithm can
    be run in either batch  or online  mode

__Down sides:__ 

-   It can be very slow

-   Need to choose the learning rate
    $\gamma_t$
---

## Other algorithms

There are many  variations on the fitting algorithm

__Stochastic gradient descent__: (SGD) discussed in the optimization lecture

The rest are variations that use lots of tricks

* RMSprop
* Adam
* Adadelta
* Adagrad
* Adamax
* Nadam
* Ftrl

---

## Regularizing neural networks

NNets can almost always achieve 0 training error. Even with regularization

Flavors:

-   a complexity penalization term

-   early stopping on the back propagation algorithm used for fitting


Explicit regularization $\longrightarrow$ solve $\min \hat{R} + \rho(\alpha,\beta)$

-   __Weight decay:__  This is like
    ridge regression in that we penalize the squared Euclidean norm of
    the weights $\rho(\alpha,\beta) = \sum \beta^2 + \sum \alpha^2$

-   __Weight elimination:__  This
    encourages more shrinking of small weights
    $\rho(\alpha,\beta) =  \sum \frac{\beta^2}{1+\beta^2} + \sum \frac{\alpha^2}{1 + \alpha^2}$ or Lasso-type


__Dropout:__

In each epoch, randomly choose $z\%$ of the nodes and set the weights to zero.

---

## Other common pitfalls

There are three areas to watch out for

- __Nonconvexity:__  The neural network optimization problem is non-convex. This makes any numerical solution highly dependent on the initial values. These must be

  - chosen carefully, typically random near 0. .hand[DON'T] use all 0.

  - regenerated several times to check sensitivity

-   __Scaling:__  Be sure to
    standardize the covariates before training

-   __Number of hidden units:__  It is generally
    better to have too many hidden units than too few (regularization
    can eliminate some).


- __Sifting the output:__ 
    
    -   Choose the solution that minimizes training
    error

    -   Choose the solution that minimizes the
    penalized  training error

    -   Average the solutions across runs

---

## Tuning parameters

There are many.

* Regularization
* Stopping criterion
* learning rate
* Architecture
* Dropout %
* others...

These are hard to tune.

Often choose "some" with a validation set.

--

More often, people set them all arbitrarily

---

## Thoughts on NNets

Off the top of my head, without lots of justification

.pull-left[

__Why don't statisticians like them?__

.center[.larger[`r fa("angry", fill=red)`]]

- There is almost no theory (certainly nothing for deep nets)
- Theory is (as far as we can tell) nearly impossible
- Statistical theory applies to global minima, here, only local determined by the optimizer
- No good understanding of when they work
- In large part, NNets look like logistic regression + feature creation. We understand that well, and in many applications, it performs as well
- Explosion of tuning parameters without a way to decide
- Require massive datasets to work
]

.pull-right[

__Why are they hot?__

.center[.larger[`r fa("fire-alt", fill=orange)`]]

- Perform exceptionally well on typical CS tasks (images, translation)
- Take advantage of SOTA computing (parallel, GPUs)
- Very good for multinomial logistic regression
- An excellent example of "transfer learning"
- They generate pretty pictures (the nets, pseudo-responses at hidden units)

]

---

## Keras

Most people who do deep learning use Python $+$ Keras $+$ Tensorflow

It takes some work to get all this software up and running.

It is possible to do in with R using an [interface to Keras](https://keras.rstudio.com/index.html).

We did this last week, so see there.

```{r embeddings, echo=FALSE,include=FALSE,fig.width=16,fig.height=9}
library(maniTools)
num_pts = 600
d = 2   #target dimension
k = 8   #k nearest neighbors
sim_data <- swiss_roll(num_pts)
pca_dr <- sim_data$data %>% center_and_standardise() %>% prcomp()
proj_pca <- sim_data$data %*% pca_dr$rotation[,1:2] %>% scale()
proj_hess <- Hessian_LLE(sim_data$data, k = k, d = d)$projection %>% scale()
proj_LLE <- LLE2(sim_data$data, dim = d, k = k) %>% scale()
all = cbind(rbind(proj_pca, proj_LLE, proj_hess), sim_data$colors)
colnames(all) = c('x','y','col')
all = as_tibble(all)
all$method = rep(c("PCA","Hessian maps","LLE"), each=num_pts) 
ggplot(all, aes(x,y,color=col)) + geom_point() +
  scale_color_viridis_c() +
  facet_wrap(~method, scales="free") + 
  theme(legend.position = "none", 
        axis.line = element_blank(), axis.text = element_blank(),
        axis.ticks = element_blank(), axis.title = element_blank())
```

---
class: middle, center
background-image: url(rmd_gfx/23-nnets-other/embeddings-1.svg)
background-size: cover



.primary[.larger[Next time...]]

.primary[.larger[Module]] .huge-blue-number[5]

.primary[.larger[unsupervised learning]]




