---
title: "Laplacian Eigenmaps"
author: DJM
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  learnr::tutorial:
    progressive: true
    theme: journal
    highlight: kate
    ace_theme: solarized_dark
    includes:
      in_header: !expr system.file("tutorials/tutorials-css.html",package="UBCstat406labs")
runtime: shiny_prerendered
---

<!--
Derived from IU Stat 432 week9.Rmd
css: "/css/tutorials.css"
-->

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(knitr)

tutorial_options(exercise.timelimit = 5, exercise.checker = gradethis::grade_learnr)

knitr::opts_chunk$set(echo = FALSE)
```


## Laplacian Eigenmaps 

### Generate the Data

Run the following code that generates a set of points in 2-dimension space.  You should see what appears to be a smaller circle inside a larger circle.

```{r circlesdata, exercise=TRUE, warning=FALSE, message=FALSE}
library(tidyverse)
tt = seq(0,2*pi,len=50)
tt2 = seq(0,2*pi,len=75)
c1 = data.frame(x=cos(tt),y=sin(tt),grp=1)
c2 = data.frame(x=1.5*cos(tt2),y=1.5*sin(tt2),grp=2)
circles = bind_rows(c1,c2)
n = nrow(circles)
ggplot(circles,aes(x=x,y=y,color=as.factor(grp))) + geom_point() + 
   theme_minimal(base_family = 'serif') + theme(legend.position = 'none')
```

### K-means

The goal is to identify the inner points from the outer points.  Let's perform K means with $K=2$. Plot your clusters.

```{r kmeans-setup, warning=FALSE, message=FALSE}
library(tidyverse)
tt = seq(0,2*pi,len=50)
tt2 = seq(0,2*pi,len=75)
c1 = data.frame(x=cos(tt),y=sin(tt),grp=1)
c2 = data.frame(x=1.5*cos(tt2),y=1.5*sin(tt2),grp=2)
circles = bind_rows(c1,c2)
n = nrow(circles)
```

```{r kmeans, exercise=TRUE}
raw_kmeans = kmeans(circles[,1:2], 2, nstart = 10)
circles$kmeans = raw_kmeans$cluster
ggplot(circles,aes(x=x,y=y,color=as.factor(kmeans))) + geom_point() + 
   theme_minimal(base_family = 'serif') + theme(legend.position = 'none')
```

How well did you do?

### Laplacian Eigenmaps

Use Laplacian Eigenmaps with Gaussian Kernel and $\gamma=0.01$. Use two eigenvectors. Plot the data in the new coordinate system and color by the original groups. (I believe you can use the last 2 eigenvectors here.)

```{r kpca-setup, warning=FALSE, message=FALSE}
library(tidyverse)
tt = seq(0,2*pi,len=50)
tt2 = seq(0,2*pi,len=75)
c1 = data.frame(x=cos(tt),y=sin(tt),grp=1)
c2 = data.frame(x=1.5*cos(tt2),y=1.5*sin(tt2),grp=2)
circles = bind_rows(c1,c2)
n = nrow(circles)
raw_kmeans = kmeans(circles[,1:2], 2, nstart = 10)
circles$kmeans = raw_kmeans$cluster
```

```{r kpca, exercise=TRUE, warning=FALSE, message=FALSE}
Delta = as.matrix(dist(as.matrix(scale(circles[,1:2]))))
K = exp(-Delta^2/.01)
L = diag(n) - diag(1/rowSums(K)) %*% K
EE = eigen(L, symmetric = TRUE)
KernCoords = EE$vectors[,124:125] %*% diag(EE$values[124:125])
LapEigenmaps = tibble(x=KernCoords[,1], y=KernCoords[,2],grp = circles$grp)
ggplot(LapEigenmaps, aes(x=x,y=y,color=as.factor(grp))) + geom_point() + 
   theme_minimal(base_family = 'serif') + theme(legend.position = 'none') 
```

### K-means on Laplacian Eigenmaps

Now apply kmeans to the data in the new coordinate system. 

```{r kernelkmeans-setup, warning=FALSE,message=FALSE}
library(tidyverse)
tt = seq(0,2*pi,len=50)
tt2 = seq(0,2*pi,len=75)
c1 = data.frame(x=cos(tt),y=sin(tt),grp=1)
c2 = data.frame(x=1.5*cos(tt2),y=1.5*sin(tt2),grp=2)
circles = bind_rows(c1,c2)
n = nrow(circles)
raw_kmeans = kmeans(circles[,1:2], 2, nstart = 10)
circles$kmeans = raw_kmeans$cluster
Delta = as.matrix(dist(as.matrix(scale(circles[,1:2]))))
K = exp(-Delta^2/.01)
L = diag(n) - diag(1/rowSums(K)) %*% K
EE = eigen(L, symmetric = TRUE)
KernCoords = EE$vectors[,124:125] %*% diag(EE$values[124:125])
LapEigenmaps = tibble(x=KernCoords[,1], y=KernCoords[,2],grp = circles$grp)
```

```{r kernelkmeans, exercise=TRUE}
circles$kern_kmeans = kmeans(KernCoords, 2, nstart = 10)$cluster
ggplot(circles, aes(x=x,y=y,color=as.factor(kern_kmeans))) + geom_point() + 
   theme_minimal(base_family = 'serif') + theme(legend.position = 'none')
```

How well did you do?

